---
title: "Procedures to integrate the identified metabolites using deconvoluted data in four algorithms (MZmine, MS-DIAL, eRah, and MSHub) and against libraries' spectra of NIST 20 (**mainlib** and **replib**) using NIS MS Search software - Rtx-5 Semi-standard non-polar colum - SH-Rtx-Wax Polar colum"
author: "Jefferson Pastuna"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
usethis::git_vaccinate()

```

## Loading libraries and data

Loading the necessary libraries.

```{r }

library(readxl)
library(dplyr)
library(stringr)
library(purrr)
library(ggplot2)
library(ComplexHeatmap)
library(patchwork)

```

Loading the identified list of each software, MZmine, MS-DIAL, eRah, and MSHub.

```{r }

# Loading features list of each software with metabolite identification
mzmine_flist <- read_excel(
  "../X_hylaeae_metabolomics/Result/MZmine_GCMS_polar_IDs.xlsx",
  col_types = "text")
msdial_flist <- read_excel(
  "../X_hylaeae_metabolomics/Result/MSDIAL_GCMS_polar_IDs.xlsx",
  col_types = "text")
erah_flist <- read_excel(
  "../X_hylaeae_metabolomics/Result/eRah_GCMS_polar_IDs.xlsx",
  col_types = "text")
mshub_flist <- read_excel(
  "../X_hylaeae_metabolomics/Result/MSHub_GCMS_polar_IDs.xlsx",
  col_types = "text")
# Merge all feature list
all_flist <- bind_rows(
  mzmine_flist %>% mutate(Alignment_ID = as.character(Alignment_ID),
                          Software = "MZmine"),
  msdial_flist %>% mutate(Alignment_ID = as.character(Alignment_ID),
                          Software = "MS-DIAL"),
  erah_flist %>% mutate(Alignment_ID = as.character(Alignment_ID),
                        Software = "eRah"),
  mshub_flist %>% mutate(Alignment_ID = as.character(Alignment_ID),
                         Software = "MSHub"))

```

The following code extracts the top 5 match ions with the NIST library. This data is very important in the following steps: to compare features using the EI spectrum and to annotate metabolite names in the feature list for statistical analysis.

```{r }

# Function to extract the 5 most abundant ions that matched the NIST library
five_match_ion <- function(exp_mz, exp_int, nist_mz) {
  # Avoid NA values
  if (is.na(exp_mz) | is.na(exp_int) | is.na(nist_mz)) {
    return(NA_character_)
  }
  # Extracting spectrum data
  mz_exp <- as.numeric(str_split(exp_mz, ",")[[1]])
  int_exp <- as.numeric(str_split(exp_int, ",")[[1]])
  mz_nist <- as.numeric(str_split(nist_mz, ",")[[1]])
  # Ensuring the ions and intensity values
  if (length(mz_exp) != length(int_exp)) {
    return(NA_character_)
  }
  # Experimental DataFrame
  df_exp <- tibble(mz = mz_exp, intensity = int_exp)
  # Finding ions match
  df_match <- df_exp %>%
    filter(mz %in% mz_nist)
  # If there is no match
  if (nrow(df_match) == 0) {
    return(NA_character_)
  }
  # 5 most abundant ions that matched the NIST library
  df_top5 <- df_match %>%
    arrange(desc(intensity)) %>%
    slice_head(n = 5)
  # Return as string (mz)
  paste(df_top5$mz, collapse = ",")
}
# Using the function five_match_ion to extract the top 5 ions
all_flist <- all_flist %>%
  mutate(top5_match_mz = pmap_chr(list(exp_mz_ns, exp_int_ns, nist_mz),
                                  five_match_ion))

```

## Clean repeated metabolites

The repeated metabolites were cleaned using the same framework used in the independent software: if the same metabolite was assigned in two or more deconvolution software, the best match factor was prioritized in the cases where the RI values are between 0 and 10 (the in-house library was prioritized, and then the filter was applied). For metabolites with RI errors between 10 and 100, the metabolite name was assigned to the software with the lower RI error; in cases of identical RI errors, the best match factor was prioritized. In cases where metabolites had an RI error > 100 or were NA, the metabolite name was assigned to the software with the highest match factor. Metabolites that do not meet the condition described above were tagged as "Software repeated (RI error <10)", "Software repeated (RI error 10-100)", and "Software repeated (RI error >100/NA)", accordingly.

```{r }

# Filtering only features with clean IDs in each software
nflag_flist <- all_flist %>%
  filter(is.na(Match_flag))
# Filtering the rest of the features (flagged features)
yflag_flist <- all_flist %>%
  filter(!is.na(Match_flag))
# Flag repeated metabolites identified by different software
rep_ids <- nflag_flist %>%
  group_by(Name) %>%
  mutate(in_0_10   = !is.na(RI_Error) & RI_Error <= 10,
         in_10_100 = !is.na(RI_Error) & RI_Error > 10 & RI_Error <= 100,
         has_0_10   = any(in_0_10),
         has_10_100 = any(in_10_100),
         is_ikiam = Library == "ikiam_npl_gcms_lib") %>%
  # Define active group
  mutate(active_group = case_when(
    has_0_10 ~ "0_10",
    !has_0_10 & has_10_100 ~ "10_100",
    TRUE ~ "gt100")) %>%
  # Define who participates in the ranking
  mutate(candidate = case_when(active_group == "0_10"   ~ in_0_10,
                               active_group == "10_100" ~ in_10_100,
                               TRUE ~ TRUE)) %>%
  # Order for candidates only
  arrange(Name,
    desc(candidate),  # Candidates first
    # PRIORITY 1
    desc(ifelse(candidate & active_group == "0_10", is_ikiam, FALSE)),
    desc(ifelse(candidate & active_group == "0_10", MatchFactor, NA)),
    ifelse(candidate & active_group == "0_10", RI_Error, Inf),
    # PRIORITY 2
    ifelse(candidate & active_group == "10_100", RI_Error, Inf),
    desc(ifelse(candidate & active_group == "10_100", MatchFactor, NA)),
    # PRIORITY 3
    desc(ifelse(candidate & active_group == "gt100", MatchFactor, NA))) %>%
  group_by(Name) %>%
  mutate(rank_name = ifelse(candidate, row_number(), NA_integer_)) %>%
  ungroup() %>%
  # Labeled
  mutate(Match_flag = case_when(
    rank_name == 1 ~ Match_flag,
    active_group == "0_10" ~ "Software repeated (RI error <10)",
    active_group == "10_100" ~ "Software repeated (RI error 10-100)",
    TRUE ~ "Software repeated (RI error >100/NA)")) %>%
  select(-in_0_10, -in_10_100, -has_0_10, -has_10_100,
         -is_ikiam, -active_group, -candidate, -rank_name)
# Reintegrate the flagged features list with the no-flagged-features list
rep_ids <- bind_rows(rep_ids, yflag_flist)

```

This output is the first result of the cleaning step of the feature list of metabolites identified using multiple deconvolution platforms. The result is a feature list with no duplicate metabolites. Only the metabolites with the best match, based on experimental data from the deconvolution software, are retained; the rest are tagged as repeated metabolites.

## Discrepancies between software

Discrepancies between software refer to the "same deconvoluted feature" identified under different metabolite names across different deconvolution software. To determine if the same feature was identified with different compound names across different deconvolution software, the experimental top 5 ions (more abundant) that matched with the NIST library will be used, and the feature in a short retention time (we use 0.04 min, this time was determined experimentally, since the time shift observed in the raw data) will be compared with each other.

### Group by RT

The following RT grouping algoritm, first group features in chain from minor RT from mayor RT, if the close feature are in the RT tolerance, in this case 0.04 min, its gruped in the same group; in the second stage, the algoritm separate feature from the group if it is out of RT toletance by comparing the feature with lowest RT and the rest of the feature in the group (this break the chain effec); them the firts group method will be apply again to group the separated features in a new group, first and secong grouping algoritm is repeated until the group have not feature out of the RT tolerence.

```{r }

# Filtering features without a flag
nflag_rep_ids <- rep_ids %>%
  filter(is.na(Match_flag)) %>%
  mutate(RT = as.numeric(RT))
# Filtering the rest of the features
yflag_rep_ids <- rep_ids %>%
  filter(!is.na(Match_flag)) %>%
  mutate(RT = as.numeric(RT))
# Function to group the features with close RT
rt_group <- function(df, tol = 0.04, max_iter = 20) {
  iter <- 1
  # Beginning an empty group
  df <- df %>% arrange(RT) %>% mutate(RT_group = NA_integer_)
  repeat {
    message("Iteración: ", iter)
    # Start grouping
    df <- df %>%
      arrange(RT) %>%
      mutate(RT_group = {
        grp <- integer(n())
        current_group <- 1
        anchor <- RT[1]
        grp[1] <- current_group
        for (i in 2:n()) {
          if ((RT[i] - anchor) <= tol) {
            grp[i] <- current_group
            } else {
              current_group <- current_group + 1
              anchor <- RT[i]
              grp[i] <- current_group
            }}
        grp
        })
    # Clean group ensuring RT tolerance
    df <- df %>%
      group_by(RT_group) %>%
      mutate(RT_min = min(RT),
             group_again = ifelse(RT - RT_min > tol,
                                  "group_again",
                                  NA_character_)) %>%
      ungroup()
    # Condition to stop grouping
    if (!any(df$group_again == "group_again", na.rm = TRUE) || iter > max_iter) {
      message("Finalizado en iteración: ", iter)
      break}
    # Clean RT_group if it corresponds to another group
    df <- df %>%
      mutate(RT_group = ifelse(group_again == "group_again", NA, RT_group)) %>%
      arrange(RT)
    iter <- iter + 1
  }
  # Delete auxiliary columns
  df %>%
    select(-RT_min, -group_again)
}
# Grouping the features with a short RT window (0.04 min)
rep_rt <- rt_group(df = nflag_rep_ids, tol = 0.04, max_iter = 20) %>%
  # Measure the group amount
  group_by(RT_group) %>%
  mutate(n_in_group = n()) %>%
  # Replace the NA value in the groups with one feature
  mutate(RT_group = ifelse(n_in_group == 1, NA, RT_group)) %>%
  ungroup() %>%
  select(-n_in_group)
# Reintegrate the flagged features list with the no-flagged-features list
rep_rt  <- bind_rows(rep_rt, yflag_rep_ids)

```

### Group by EI spectrum

The following code compare the top 5 ions of features in the same group, the code determine what ion is most frecuency in the group, and use that number to find another ions that have same frecuency, if some feature do not have comon ion the code add a "group_again" tag (this feature is exclude of the group because could be other metabolite that elute in close RT). In other words, the feature that shares the same ions could correspond to a single feature, but it matched with different metabolites due to the use of 4 different deconvolution algorithms.

**NOTE.** We use the top 5 ions to compare features rather than the complete EI-deconvoluted mass spectrum because, for the statistical analysis, we will use the most abundant ions of the mass spectrum to annotate the metabolite name. Also, if two features coelute at the same or very close retention time, we can elucidate them or consider them as two metabolites only if their mass spectra differ (i.e., have unique ions for each compound). Therefore, the code seeks the characteristic ions of possible coeluting compounds, and these ions would be among the first 5 ions matched to the spectrum library.

```{r }

# Filtering features without a flag
nflag_rep_ids <- rep_rt %>%
  filter(is.na(Match_flag))
# Filtering the rest of the features
yflag_rep_ids <- rep_rt %>%
  filter(!is.na(Match_flag))
# Function to extract a numeric vector of the EI spectrum
parse_top5_mz <- function(x) {
  if (is.na(x)) return(numeric(0))
  as.numeric(str_split(x, ",")[[1]])
}
# Search shared ions between groups
share_ion <- nflag_rep_ids %>%
  mutate(mz_list = map(top5_match_mz, parse_top5_mz)) %>%
  group_by(RT_group) %>%
  group_modify(~ {
    data <- .x
    key  <- .y
    n_group <- nrow(data)
    # Case: NA or group of 1
    if (is.na(key$RT_group) || n_group <= 1) {
      data$Share_ion <- NA_character_
      return(data)
    }
    # Count ion frequency
    freq <- table(unlist(data$mz_list))
    # Case: large groups without shared ions
    if (n_group >= 3 && max(freq) == 1) {
      data$Share_ion <- "group_again"
      return(data)
    }
    # Logic according to group size
    if (n_group == 2) {
      # Only ions shared in both features
      top_ions <- as.numeric(names(freq[freq == 2]))
    } else {
        # Original logic (maximum frequency)
      max_freq <- max(freq)
      top_ions <- as.numeric(names(freq[freq == max_freq]))
    }
    # Row assignment
    data$Share_ion <- map_chr(data$mz_list, ~ {
      if (length(top_ions) > 0 && any(top_ions %in% .x)) {
        paste(intersect(.x, top_ions), collapse = ",")
      } else {
        "group_again"
      }
    })
    data
  }) %>%
  ungroup() %>%
  select(-mz_list)
# Reintegrate the flagged features list with the no-flagged-features list
share_ion  <- bind_rows(share_ion, yflag_rep_ids)

```

In the first stage, we grouped features with a short RT (we used an RT tolerance of 0.04 min) and compared the 5 most abundant ions matching the NIST library with other features in the same group. If the EI spectrum of a feature from the group did not share ions with the group, it is separated from the group (these features will be analyzed post hoc). Now we will use the groups to select the best match within each group (since each group shares ions with the others, we can treat each group as a feature with multiple matches).

```{r }

# Filtering features with group and without "group_again" flag
gp_share_ion <- share_ion %>%
  filter(!(Share_ion == "group_again") & !is.na(Share_ion))
# Filtering the rest of features
rest_share_ion <- share_ion %>%
  filter(Share_ion == "group_again" | is.na(Share_ion))
# Flag RT repeated metabolites
flag_share_ion <- gp_share_ion %>%
  group_by(RT_group) %>%
  mutate(in_0_10   = !is.na(RI_Error) & RI_Error <= 10,
         in_10_100 = !is.na(RI_Error) & RI_Error > 10 & RI_Error <= 100,
         has_0_10   = any(in_0_10),
         has_10_100 = any(in_10_100),
         is_ikiam = Library == "ikiam_npl_gcms_lib") %>%
  # Define active group
  mutate(active_group = case_when(
    has_0_10 ~ "0_10",
    !has_0_10 & has_10_100 ~ "10_100",
    TRUE ~ "gt100")) %>%
  # Define who participates in the ranking
  mutate(candidate = case_when(active_group == "0_10"   ~ in_0_10,
                               active_group == "10_100" ~ in_10_100,
                               TRUE ~ TRUE)) %>%
  # Order for candidates only
  arrange(RT_group,
    desc(candidate),  # Candidates first
    # PRIORITY 1
    desc(ifelse(candidate & active_group == "0_10", is_ikiam, FALSE)),
    desc(ifelse(candidate & active_group == "0_10", MatchFactor, NA)),
    ifelse(candidate & active_group == "0_10", RI_Error, Inf),
    # PRIORITY 2
    ifelse(candidate & active_group == "10_100", RI_Error, Inf),
    desc(ifelse(candidate & active_group == "10_100", MatchFactor, NA)),
    # PRIORITY 3
    desc(ifelse(candidate & active_group == "gt100", MatchFactor, NA))) %>%
  group_by(RT_group) %>%
  mutate(rank_name = ifelse(candidate, row_number(), NA_integer_)) %>%
  ungroup() %>%
  # Labeled
  mutate(Match_flag = case_when(
    rank_name == 1 ~ Match_flag,
    active_group == "0_10" ~ "RT repeated (RI error <10)",
    active_group == "10_100" ~ "RT repeated (RI error 10-100)",
    TRUE ~ "RT repeated (RI error >100/NA)")) %>%
  select(-in_0_10, -in_10_100, -has_0_10, -has_10_100,
         -is_ikiam, -active_group, -candidate, -rank_name)
# Reintegrate all data
rep_share_ion <- bind_rows(flag_share_ion, rest_share_ion)

```

In the following code we will filter the features flagged in the Match_flag column, and them we will analysed the features with the tag "group_again". Also, we will add a tag in the column Match_flag if there are more of 1 feature in en the group (that mean that are features in close RT, 0.04 in this case, that was deconvoluted as different compound).

The code aimed to verify whether the features flagged as "group_again" could be joined to the existing group. The code ordering the RT from minor to major, them search neigbord features in the ± RT tolerance, if there are feature in the RT tolerance (we use a windows of ± 0.04 min) the code verified if the feature share the top 5 ions with the caracteristics ions of the group (ions of the columm Share_ion), if there are share ion, the code register the ions in the Share_ion column, in contrast, if there are not share ion, the feature will be tag as "Coelute_compound".

Also, a new column, "Matched_RT_group," was added to record the group that the "group_again"- flagged feature joins. This new column will be used to verify whether a new group will be created. If a new group is created, the last code needs to be applied again to select the best match. On the other hand, if no new group is created, it means the consensus identification list is ready.

```{r }

# Filtering features without repeated feature in the short RT
n_rep_share_ion <- rep_share_ion %>%
  filter(is.na(Match_flag))
# Filtering the rest of features
y_rep_share_ion <- rep_share_ion %>%
  filter(!is.na(Match_flag))
# Function to extract numeric vector of EI spectrum
parse_ei_mz <- function(x) {
  if (is.na(x)) return(numeric(0))
  as.numeric(str_split(x, ",")[[1]])
}
# Function to scan coelution compounds
scan_coelution <- function(df, rt_tol = 0.04) {
  df <- df %>%
    arrange(RT) %>%
    mutate(mz_top5  = map(top5_match_mz, parse_ei_mz),
           mz_share = map(
             Share_ion,
             ~ ifelse(.x == "group_again" | is.na(.x), NA, .x)) %>%
             map(parse_ei_mz),
           # New column
           Matched_RT_group = NA_integer_)
  n <- nrow(df)
  for (i in seq_len(n)) {
    # Only work with group_again
    if (is.na(df$Share_ion[i]) || df$Share_ion[i] != "group_again") next
    rt_i  <- df$RT[i]
    mz_i  <- df$mz_top5[[i]]
    grp_i <- df$RT_group[i]
    found_match   <- FALSE
    matched_ions  <- c()
    matched_group <- NA_integer_
    # Search back
    j <- i - 1
    while (j >= 1) {
      # Stop if RT tolerance is exceeded
      if ((rt_i - df$RT[j]) > rt_tol) break
      # Exclude same RT_group and NA
      if (!is.na(df$RT_group[j]) &&
          df$RT_group[j] != grp_i) {
        ions_j <- df$mz_share[[j]]
        if (length(ions_j) > 0) {
          inter <- intersect(mz_i, ions_j)
          # Match found
          if (length(inter) > 0) {
            matched_ions <- unique(c(matched_ions, inter))
            matched_group <- df$RT_group[j]
            found_match <- TRUE
          }
        }
      }
      j <- j - 1
    }
    # Search forward
    j <- i + 1
    while (j <= n) {
      # Stop if RT tolerance is exceeded
      if ((df$RT[j] - rt_i) > rt_tol) break
      if (!is.na(df$RT_group[j]) &&
          df$RT_group[j] != grp_i) {
        ions_j <- df$mz_share[[j]]
        if (length(ions_j) > 0) {
          inter <- intersect(mz_i, ions_j)
          # Match found
          if (length(inter) > 0) {
            matched_ions <- unique(c(matched_ions, inter))
            matched_group <- df$RT_group[j]
            found_match <- TRUE
          }
        }
      }
      j <- j + 1
    }
    # Final assignment
    if (found_match) {
      df$Share_ion[i] <- paste(sort(matched_ions),
                               collapse = ",")
      df$Matched_RT_group[i] <- matched_group
    } else {
      df$Share_ion[i] <- "Coelute_compound"
    }
  }
  df %>%
    select(-mz_top5, -mz_share)
}
# Determine if there are coelution compounds
coelu_scan <- scan_coelution(n_rep_share_ion, rt_tol = 0.04)

```

In the next code snippet, we verify whether a "group_again"-flagged feature joins a new group; the code checks whether there are any values in the Matched_RT_group column.

```{r }

# Verify if new groups were created
if (all(is.na(coelu_scan$Matched_RT_group))) {
  
  message("Successful! The final identification list is ready")
  
} else {
  
  message("New groups were created, repeat the last code")
}

```

Because the "group_again"-flagged features do not create a new group (as shown above), we can extract key summaries from the final consensus identification list. Some of the important results that we can report are (a) metabolite identification between different deconvolution software (agreement between software), (b) metabolites selected as best but with identification discrepancies between different deconvolution software, (c) coeluted features (different features that eluted in the close RT, in this case we used 0.04 min), and (d) the consensus identification list.

In the following code, we merge all feature lists from four different deconvolution software.

```{r }

# Flagging coeluted features
coelu_scan <- coelu_scan %>%
  mutate(Match_flag = case_when(
    Share_ion == "Coelute_compound" ~ "Coelute IDs"))
# Reintegrating all feature list resulted from rep_share_ion
coelu_feat <- bind_rows(coelu_scan, y_rep_share_ion)

```

## Result

To report (a) metabolite identification between different deconvolution software (agreement between software), we filter the final metabolite names and search if that metabolite was identified in other deconvolution software; it will be reported as metabolites identified in the four deconvolution software, those identified in 3 deconvolution software, 2 deconvolution software, and those identified in a specific deconvolution software..

```{r }

# Parameters
rt_tol <- 0.05
# 1. Consensus metabolite list
id_best <- coelu_feat %>%
  filter(is.na(Match_flag) |
           Match_flag == "" |
           Match_flag == "Coelute IDs") %>%
  arrange(Name, RT)
# 2. Use consensus names to search in Original dataframe
consensus_names <- unique(id_best$Name)
candidate_rows <- coelu_feat %>%
  filter(Name %in% consensus_names)
# 3. Compare RT consensus vs candidate rows (Keep only rows within ±0.05 RT)
id_agre <- map_dfr(seq_len(nrow(id_best)), function(i) {
  # Consensus feature
  ref_name <- id_best$Name[i]
  ref_rt   <- id_best$RT[i]
  # Search same metabolite name
  candidates <- candidate_rows %>%
    filter(Name == ref_name)
  # RT filtering
  matched <- candidates %>%
    filter(abs(RT - ref_rt) <= rt_tol) %>%
    mutate(Consensus_RT = ref_rt)
  matched
})
# Remove duplicates if necessary
id_agre <- id_agre %>%
  distinct()
# 4. Count software occurrence
id_software <- id_agre %>%
  group_by(Name) %>%
  summarise(n_softwares = n_distinct(Software),
            softwares = paste(
              sort(unique(Software)),
              collapse = ", "),
            .groups = "drop")
# 5. Software classification
id_classify <- id_software %>%
  mutate(Software_class = case_when(
    n_softwares == 4 ~ "Identified in 4 softwares",
    n_softwares == 3 ~ "Identified in 3 softwares",
    n_softwares == 2 ~ "Identified in 2 softwares",
    n_softwares == 1 & softwares == "MZmine" ~ "Unique to MZmine",
    n_softwares == 1 & softwares == "MS-DIAL" ~ "Unique to MS-DIAL",
    n_softwares == 1 & softwares == "eRah" ~ "Unique to eRah",
    n_softwares == 1 & softwares == "MSHub" ~ "Unique to MSHub",
    TRUE ~ "Other")) %>%
  # Adding column to UpSet plot
  mutate(MZmine = ifelse(str_detect(softwares, "\\bMZmine\\b"), 1, 0),
         `MS-DIAL` = ifelse(str_detect(softwares, "\\bMS-DIAL\\b"), 1, 0),
         MSHub = ifelse(str_detect(softwares, "\\bMSHub\\b"), 1, 0),
         eRah = ifelse(str_detect(softwares, "\\beRah\\b"), 1, 0))
# 6. Final output
soft_classify_feat <- id_best %>%
  left_join(id_classify,
            by = "Name")

```

### Identification by software

To demonstrate how the metabolites were identified across different software, we will present the results in an UpSet plot. We will inspect how the software identified at all identification levels, at levels 1 and 2 only, and at level 3 only. It helps to know which software is better in case the searcher would like to use only one.

```{r echo=TRUE, fig.width = 10, fig.height = 3}

# All identification levels
all_upset_plot <- soft_classify_feat %>%
  select(Name, MZmine, 'MS-DIAL', MSHub, eRah)
# Identification level 1 and 2
il12_upset_plot <- soft_classify_feat %>%
  filter(IL == "Level 1" | IL == "Level 2") %>%
  select(Name, MZmine, 'MS-DIAL', MSHub, eRah)
# Identification level 3
il3_upset_plot <- soft_classify_feat %>%
  filter(IL == "Level 3") %>%
  select(Name, MZmine, 'MS-DIAL', MSHub, eRah)
# Make the combination matrix
all_comb_mat = make_comb_mat(all_upset_plot)
il12_comb_mat = make_comb_mat(il12_upset_plot)
il3_comb_mat = make_comb_mat(il3_upset_plot)
# UpSet plot
all_upset <- UpSet(all_comb_mat, row_names_gp = gpar(fontsize = 10),
                   column_title = "Level 1 to 3 metabolites",
                   top_annotation =
                     upset_top_annotation(all_comb_mat,
                                          gp = gpar(fill = "#F8766D",
                                                    col = "#F8766D"),
                                          add_numbers = TRUE,
                                          annotation_name_rot = 90),
                   right_annotation =
                     upset_right_annotation(all_comb_mat,
                                            gp = gpar(fill = "#1C6AA8",
                                                      col = "#1C6AA8"),
                                            add_numbers = TRUE),
                   set_order = c("MZmine", "MS-DIAL", "eRah", "MSHub"))
il12_upset <- UpSet(il12_comb_mat, row_names_gp = gpar(fontsize = 10),
                    column_title = "Level 1 and 2 metabolites",
                    top_annotation =
                      upset_top_annotation(il12_comb_mat,
                                           gp = gpar(fill = "#F8766D",
                                                     col = "#F8766D"),
                                           add_numbers = TRUE,
                                           annotation_name_rot = 90),
                    right_annotation =
                      upset_right_annotation(il12_comb_mat,
                                             gp = gpar(fill = "#1C6AA8",
                                                       col = "#1C6AA8"),
                                             add_numbers = TRUE),
                    set_order = c("MZmine", "MS-DIAL", "eRah", "MSHub"))
il3_upset <- UpSet(il3_comb_mat, row_names_gp = gpar(fontsize = 10),
                   column_title = "Level 3 metabolites",
                   top_annotation =
                     upset_top_annotation(il3_comb_mat,
                                          gp = gpar(fill = "#F8766D",
                                                    col = "#F8766D"),
                                          add_numbers = TRUE,
                                          annotation_name_rot = 90),
                   right_annotation =
                     upset_right_annotation(il3_comb_mat,
                                            gp = gpar(fill = "#1C6AA8",
                                                      col = "#1C6AA8"),
                                            add_numbers = TRUE),
                   set_order = c("MZmine", "MS-DIAL", "eRah", "MSHub"))
all_upset + il12_upset + il3_upset

```

### Coeluted compounds

To report (b) metabolites selected as the best but with identification discrepancies between deconvolution software, we define the meaning of discrepancies; discrepancies between software will be considered when features with a short retention time (e.g., 0.04 min) and spectral similarity (at least one ion match within the group) are annotated with different metabolite names. In the following code, we plot the mass spectrum of the features in each group, which helps us visually inspect the results of the code filtering.

```{r echo=TRUE, warning=FALSE}

# Function to parse numeric vectors
parse_num <- function(x) {
  if (is.na(x) || x == "") return(numeric(0))
  as.numeric(str_split(x, ",")[[1]])
}
# Function to generate EI spectrum plots by RT group
plot_gp_ei <- function(df,
                       save_pdf = FALSE,
                       outdir = NULL,
                       pdf_width = 12,
                       height_per_feature = 2.0) {
  # Create output directory if requested
  if (save_pdf) {
    if (is.null(outdir)) {
      stop("Please provide 'outdir' to save PDF files.")
    }
    if (!dir.exists(outdir)) {
      dir.create(outdir, recursive = TRUE)
    }
  }
  # Keep only grouped features
  df_plot <- df %>%
    filter(!is.na(RT_group)) %>%
    # NA as best match, others as repeated
    mutate(
      Match_flag = case_when(
        Match_flag == "RT repeated (RI error <10)" |
          Match_flag == "RT repeated (RI error 10-100)" |
          Match_flag == "RT repeated (RI error >100/NA)" ~ "Repeated",
        is.na(Match_flag) ~ "Best match",
        Match_flag == "Coelute IDs" ~ "Coelution",
        TRUE ~ Match_flag
      )
    )
  # Unique groups
  groups <- unique(df_plot$RT_group)
  # Store plots
  all_group_plots <- list()
  # Loop groups
  for (g in groups) {
    # Subset group
    group_df <- df_plot %>%
      filter(RT_group == g) %>%
      arrange(RT)
    # Shared X axis across group
    all_mz_group <- group_df$exp_mz_ns %>%
      map(parse_num) %>%
      unlist()
    x_max <- ceiling(max(all_mz_group, na.rm = TRUE))
    # Store feature plots
    feature_plots <- list()
    # Loop features
    for (i in seq_len(nrow(group_df))) {
      row <- group_df[i, ]
      # Parse spectra
      mz  <- parse_num(row$exp_mz_ns)
      int <- parse_num(row$exp_int_ns)
      # Shared ions
      shared <- parse_num(
        ifelse(row$Share_ion %in% c("Coelute IDs", "group_again", NA),
               "",
               row$Share_ion))
      # Five most abundant ion
      top5_ion_name <- parse_num(
        ifelse(row$top5_match_mz %in% NA,
               "",
               row$top5_match_mz))
      # Build spectrum dataframe
      spec_df <- data.frame(mz = mz, intensity = int)
      # Ion classification
      spec_df <- spec_df %>%
        mutate(ion_type = ifelse(mz %in% shared, "shared", "unique"),
               Share_ion = ifelse(mz %in% shared, as.character(mz),
                                  NA_character_),
               name_ion = ifelse(mz %in% top5_ion_name, as.character(mz),
                                 NA_character_))
      # Annotation text
      ann_text <- paste0("Group: ", g,
                         "\nSoftware: ", row$Software,
                         "\nRT: ", round(row$RT, 3), " min",
                         "\nMatch: ", row$MatchFactor,
                         "\nRI error: ", row$RI_Error,
                         "\n", row$Match_flag)
      # Spectrum plot
      p <- ggplot(spec_df, aes(x = mz, y = intensity, color = ion_type)) +
        geom_segment(aes(xend = mz, yend = 0), linewidth = 0.8) +
        scale_color_manual(values = c("shared" = "red", "unique" = "blue")) +
        geom_text( aes(label = name_ion), vjust = -0.2, size = 3) +
        # Shared X axis
        scale_x_continuous(limits = c(50, x_max + 2),
                           breaks = seq(50, x_max, by = 20)) +
        labs(title = NULL, x = NULL, y = "Relative abundance") +
        # Metabolite name
        annotate("text", x = 50,  y = 130, label = row$Name, hjust = 0,
                 vjust = 1, size = 4, fontface = "plain") +
        # Metadata
        annotate("text", x = x_max + 2, y = 130, label = ann_text, hjust = 0,
                 vjust = 1, size = 3) +
        theme_classic() +
        theme(legend.position = "none", plot.margin = margin(10, 90, 10, 10)) +
        coord_cartesian(clip = "off")
      feature_plots[[i]] <- p
    }
    # Vertical stack
    group_patch <- wrap_plots(feature_plots, ncol = 1) +
      plot_annotation(caption = "m/z",
                      theme = theme(
                        plot.caption = element_text(hjust = 0.5, face = "italic")))
    # Store plot
    all_group_plots[[paste0("Group_", g)]] <- group_patch
    # Save PDF
    if (save_pdf) {
      pdf_height <- max(4, nrow(group_df) * height_per_feature)
      ggsave(filename = file.path(outdir,  paste0("Group_", g, ".pdf")),
             plot = group_patch, width = pdf_width, height = pdf_height,
             units = "in")
    }
    }
  return(all_group_plots)
}
# Export plot
plot_gp <- plot_gp_ei(
  df = coelu_feat, save_pdf = TRUE,
  outdir = "../X_hylaeae_metabolomics/Result/NIST_MS_Search_polar/Group_plots")
# Example of plot
plot_gp[["Group_43"]]

```

### Agreement and discrepancies

To a graphical representation of: (b) metabolites selected as best but with identification discrepancies between different deconvolution software, (c) coeluted features (different features that eluted in the close RT, in this case we used 0.04 min). We use a pie chart, as shown below, to better represent the result.

```{r }

# Identification level 1 and 2
il12_pie_plot <- soft_classify_feat %>%
  filter(IL == "Level 1" | IL == "Level 2")
# Identification level 3
il3_pie_plot <- soft_classify_feat %>%
  filter(IL == "Level 3")
# 1. Coeluted compounds
all_coeluted <- soft_classify_feat %>%
  filter(Share_ion == "Coelute_compound") %>%
  nrow()
il12_coeluted <- il12_pie_plot %>%
  filter(Share_ion == "Coelute_compound") %>%
  nrow()
il3_coeluted <- il3_pie_plot %>%
  filter(Share_ion == "Coelute_compound") %>%
  nrow()
# 2. Discrepancies
all_discrepancies <- soft_classify_feat %>%
  filter(!is.na(RT_group) & Share_ion != "Coelute_compound") %>%
  nrow()
il12_discrepancies <- il12_pie_plot %>%
  filter(!is.na(RT_group) & Share_ion != "Coelute_compound") %>%
  nrow()
il3_discrepancies <- il3_pie_plot %>%
  filter(!is.na(RT_group) & Share_ion != "Coelute_compound") %>%
  nrow()
# 3. Agreement
all_agreement <- soft_classify_feat %>%
  filter(is.na(RT_group)) %>%
  nrow()
il12_agreement <- il12_pie_plot %>%
  filter(is.na(RT_group)) %>%
  nrow()
il3_agreement <- il3_pie_plot %>%
  filter(is.na(RT_group)) %>%
  nrow()
# Table to pie chart
all_pie_tb <- data.frame(Category = c("Coeluted",
                                     "Discrepancies",
                                     "Agreement"),
                        Count = c(all_coeluted, all_discrepancies, all_agreement))
il12_pie_tb <- data.frame(Category = c("Coeluted",
                                     "Discrepancies",
                                     "Agreement"),
                        Count = c(il12_coeluted, il12_discrepancies, il12_agreement))
il3_pie_tb <- data.frame(Category = c("Coeluted",
                                     "Discrepancies",
                                     "Agreement"),
                        Count = c(il3_coeluted, il3_discrepancies, il3_agreement))
# Labels for pie chart
all_pie_tb <- all_pie_tb %>%
  mutate(Label = paste0(Category, "\n(n = ", Count, ")"))
il12_pie_tb <- il12_pie_tb %>%
  mutate(Label = paste0(Category, "\n(n = ", Count, ")"))
il3_pie_tb <- il3_pie_tb %>%
  mutate(Label = paste0(Category, "\n(n = ", Count, ")"))
# Pie chart
all_pie_plot <- ggplot(all_pie_tb, aes(x = "", y = Count, fill = Category)) +
  geom_col(width = 1, color = "white") +
  coord_polar(theta = "y") +
  geom_text(aes(label = Label), position = position_stack(vjust = 0.5), size = 4) +
  labs(title = "Level 1 to 3 metabolites") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5, face = "plain"), legend.position = "none")
il12_pie_plot <- ggplot(il12_pie_tb, aes(x = "", y = Count, fill = Category)) +
  geom_col(width = 1, color = "white") +
  coord_polar(theta = "y") +
  geom_text(aes(label = Label), position = position_stack(vjust = 0.5), size = 4) +
  labs(title = "Level 1 and 2 metabolites") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5, face = "plain"), legend.position = "none")
il3_pie_plot <- ggplot(il3_pie_tb, aes(x = "", y = Count, fill = Category)) +
  geom_col(width = 1, color = "white") +
  coord_polar(theta = "y") +
  geom_text(aes(label = Label), position = position_stack(vjust = 0.5), size = 4) +
  labs(title = "Level 3 metabolites") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5, face = "plain"), legend.position = "none")
all_pie_plot + il12_pie_plot + il3_pie_plot

```

In the following code, we export the final list of consensus metabolite identification.

```{r }

# Export result
write_xlsx(soft_classify_feat,
           "../X_hylaeae_metabolomics/Result/NIST_MS_Search_polar/Consensus_IDs_polar.xlsx")

```

