---
title: "GC-MS metabolomics of X. hylaeae - Spectral deconvolution"
author: "Jefferson Pastuna"
date: "2026-04-15"
output:
  github_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
usethis::git_vaccinate()

```

### Introduction

Write an Introduction.

### Before to start

Write a package description.

### eRah package workflow

How to star.

```{r echo=TRUE, warning=FALSE}

# eRah package installation
#install.packages('erah')
# eRah library call
library(erah)

```

Folder files

```{r echo=TRUE}

# Delete all file that are not in folders
unlink('Data/eRah_apolar_colum/*')
# Data folder path
createdt('Data/eRah_apolar_colum/')

```

New experiment.


```{r echo=TRUE}

# Loading (*.CDF) chromatograms
instrumental <- read.csv('Data/eRah_metadata_apolar_column/eRah_apolar_colum_inst.csv', sep = ";")
# Loading metadata of the chromatograms (phenotype file)
phenotype <- read.csv('Data/eRah_metadata_apolar_column/eRah_apolar_colum_pheno.csv', sep = ";")
# Merge of metadata information with chromatograms
ex <- newExp(instrumental = instrumental,
             phenotype = phenotype,
             info = 'X. hylaeae metabolomics')

```

Compound deconvolution

Parameter

```{r echo=TRUE}

# Deconvolution parameters for chromatogram analysis between 3.2 and 49.6 min
dec.par.3to49 <- setDecPar(min.peak.width = 3,
                           min.peak.height = 450,
                           noise.threshold = 45,
                           avoid.processing.mz = c(30:32),
                           analysis.time = c(3.2, 49.6))
# Deconvolution parameters for chromatogram analysis between 49.6 and 54.0 min
dec.par.49to54 <- setDecPar(min.peak.width = 3,
                           min.peak.height = 600,
                           noise.threshold = 60,
                           avoid.processing.mz = c(30:32),
                           analysis.time = c(49.6, 54.0))
# Deconvolution parameters for chromatogram analysis between 54.0 and 56.5 min
dec.par.54to56 <- setDecPar(min.peak.width = 3,
                           min.peak.height = 1000,
                           noise.threshold = 100,
                           avoid.processing.mz = c(30:32),
                           analysis.time = c(54.0, 56.5))

```

parallel processing

```{r echo=TRUE}

plan(future::multisession, workers = 5)

```

Deconvolution

```{r echo=TRUE, warning=FALSE}

# Chromatogram deconvolution between 3.2 and 49.6 min
ex.3to49 <- deconvolveComp(ex, dec.par.3to49)
# Chromatogram deconvolution between 49.6 and 54.0 min
ex.49to54 <- deconvolveComp(ex, dec.par.49to54)
# Chromatogram deconvolution between 54.0 and 56.5 min
ex.54to56 <- deconvolveComp(ex, dec.par.54to56)

```

Alignment

```{r echo=TRUE}

# Alignment parameters
ex.al.par <- setAlPar(min.spectra.cor = 0.85,
                      max.time.dist = 3,
                      mz.range = 50:500)
# Alignment of 3.2-49.6 min dataset
ex.3to49 <- alignComp(ex.3to49, alParameters = ex.al.par)
# Alignment of 49.6-54.0 min dataset
ex.49to54 <- alignComp(ex.49to54, alParameters = ex.al.par)
# Alignment of 54.0-56.5 min dataset
ex.54to56 <- alignComp(ex.54to56, alParameters = ex.al.par)

```

Missing compound recovery

```{r echo=TRUE}

# Missing compound recovery for 3.2-49.6 min dataset
ex.3to49 <- recMissComp(ex.3to49, min.samples = 2)
# Missing compound recovery for 49.6-54.0 min dataset
ex.49to54 <- recMissComp(ex.49to54, min.samples = 2)
# Missing compound recovery for 54.0-56.5 min dataset
ex.54to56 <- recMissComp(ex.54to56, min.samples = 2)

```

Exporting alignment feature list.

```{r echo=TRUE}

# Extracting alignment feature list 3.2-49.6 min
feat_list.3to49 <- ex.3to49@Results@Alignment
# Exporting alignment feature list
write.csv(feat_list.3to49,
          file = "Result/eRah_3to49min_GCMS_apolar/eRah_feat_list_3to49min.csv")
# Extracting alignment feature list 49.6-54.0 min
feat_list.49to54 <- ex.49to54@Results@Alignment
# Exporting alignment feature list
write.csv(feat_list.49to54,
          file = "Result/eRah_49to54min_GCMS_apolar/eRah_feat_list_49to54min.csv")
# Extracting alignment feature list 54.0-56.5 min
feat_list.54to56 <- ex.54to56@Results@Alignment
# Exporting alignment feature list
write.csv(feat_list.54to56,
          file = "Result/eRah_54to56min_GCMS_apolar/eRah_feat_list_54to56min.csv")

```

# Identification

Metabolite identification was by comparing all the spectra found against a reference database. The eRah default database was used to metabolite identification. Metabolite identification was improve by exporting all the spectra found (.msp) in eRah to NIST MS Search 2.4 software.

```{r echo=TRUE, warning=FALSE}

# Identification of 3.2-49.6 min dataset
peak_iden.3to49 <- identifyComp(ex.3to49,
                                id.database = mslib,
                                mz.range = NULL,
                                n.putative = 1)
# Identification of 49.6-54.0 min dataset
peak_iden.49to54 <- identifyComp(ex.49to54,
                                 id.database = mslib,
                                 mz.range = NULL,
                                 n.putative = 1)
# Identification of 54.0-56.5 min dataset
peak_iden.54to56 <- identifyComp(ex.54to56,
                                 id.database = mslib,
                                 mz.range = NULL,
                                 n.putative = 1)

```

Exporting spectra to NIST MS Search software identification with NIST-20 library.

```{r echo=TRUE}

# 3.2-49.6 min dataset
export2MSP(peak_iden.3to49,
           store.path = "Result/eRah_3to49min_GCMS_apolar",
           alg.version = 2)
# 49.6-54.0 min dataset
export2MSP(peak_iden.49to54,
           store.path = "Result/eRah_49to54min_GCMS_apolar",
           alg.version = 2)
# 54.0-56.5 min dataset
export2MSP(peak_iden.54to56,
           store.path = "Result/eRah_54to56min_GCMS_apolar",
           alg.version = 2)

```

