---
title: "Title - Statistical analysis"
author: "Jefferson Pastuña"
date: "2025-06-24"
output:
  github_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
usethis::git_vaccinate()

```

# Introduction

This R Script aims...

# Before to start

The “*notame*” package...

## Loading libraries

The “*notame*” package and other dependency packages were installed as a first step for the analysis.

```{r echo=TRUE, message=FALSE}

# Notame package installation
#if (!requireNamespace("devtools", quietly = TRUE)) {
#  install.packages("devtools")
#}
#devtools::install_github("antonvsdata/notame", ref = "v0.3.1")

# Notame library call
library(notame)

# Dependency packages installation
install_dependencies

# Other important libraries
library(dplyr)
library(stringr)
library(writexl)
library(patchwork)
library(readxl)
library(purrr)

```

## Data preparation

Raw mzXML files was deconvoluted and alingned using MZmine en four segment to correct line base caused by the temperature ramp as suggested in (last paper of MZmine). In the following code we load the exported MZmine feature list of each segment, and than all segment was marge.

**NOTE.** The "Column" and "Ion Mode" are required for dawnstring analysis using notame R package.

```{r echo=TRUE, message=FALSE}

# Reading the feature list segment
gcms_ms1_3to49 <- read.csv(
  "../X_hylaeae_metabolomics/Data/notame/GCMS_MS1_apolar_to_notame_3to49min.csv",
  sep = ",", check.names = FALSE)
gcms_ms1_49to54 <- read.csv(
  "../X_hylaeae_metabolomics/Data/notame/GCMS_MS1_apolar_to_notame_49to54min.csv",
  sep = ",", check.names = FALSE)
gcms_ms1_54to56 <- read.csv(
  "../X_hylaeae_metabolomics/Data/notame/GCMS_MS1_apolar_to_notame_54to56min.csv",
  sep = ",", check.names = FALSE)
# Clean empty columns
gcms_ms1_3to49 <- gcms_ms1_3to49[, names(gcms_ms1_3to49) != ""]
gcms_ms1_49to54 <- gcms_ms1_49to54[, names(gcms_ms1_49to54) != ""]
gcms_ms1_54to56 <- gcms_ms1_54to56[, names(gcms_ms1_54to56) != ""]
# Merge all feature list
gcms_ms1 <- bind_rows(
  gcms_ms1_3to49 %>% mutate(Segment = "3-49min", Column = "Apolar",
                            Ion_Mode = "EI"),
  gcms_ms1_49to54 %>% mutate(Segment = "49-54min", Column = "Apolar",
                            Ion_Mode = "EI"),
  gcms_ms1_54to56 %>% mutate(Segment = "54-56min", Column = "Apolar",
                            Ion_Mode = "EI"))
# Adding a global ID
gcms_ms1 <- gcms_ms1 %>%
  mutate(Global_ID = row_number()) %>%
  # Move the new columns to the beginning
  relocate(Global_ID, Segment, Column, Ion_Mode)

```

The next step is to format the data according to the notame R package for downstream analysis. To make MetaboSets, the name R package works with these three parts of the data:

- **exprs:** feature abundances across the samples
- **pheno_data:** sample information
- **feature_data:** feature information

In the following codes, we extract the required data to make MetaboSets from the feature list exported from MZmine.

### Feature information

We extract the important information from the exported MZmine feature list related to the feature information such as, mass, retention time, and column type.

```{r echo=TRUE, message=TRUE, warning=FALSE, error=FALSE}

# Extracting the feature information from the MZmine feature list
gcms_ms1_fd <- data.frame(Global_ID = gcms_ms1$Global_ID,
                          Segment = gcms_ms1$Segment,
                          MZmine_ID = gcms_ms1$`row ID`,
                          Mass = round(gcms_ms1$`row m/z`, 4),
                          Nominal_mass = round(gcms_ms1$`row m/z`, 0),
                          RT = round(gcms_ms1$`row retention time`, 3),
                          Column = gcms_ms1$Column,
                          Ion_Mode = gcms_ms1$Ion_Mode)

```

The cosensus metabolite identificarion list will be added to the deconvoluted feature all all ions without using a deconvolution software. For better grafically representation and the higth confidence level only metabolites identified at level 1 and 2 will be used.

```{r echo=TRUE, message=TRUE, warning=FALSE, error=FALSE}

# Loading the consensus identification list
il12_ids <- read_excel(
  "../X_hylaeae_metabolomics/Result/All_column_IDs.xlsx") %>%
  filter(IL == "Level 1" | IL == "Level 2") %>%
  # Change the RT column value to numeric
  mutate(RT = as.numeric(RT))
# Selecting the short name for the plot annotation purposes
# Function to clean names
clean_name <- function(x) {
  x <- as.character(x)
  # Replace NA-like values
  x[x %in% c("NA", "", " ")] <- NA
  # Remove final " #"
  x <- str_replace(x, "\\s+#$", "")
  # Trim spaces
  x <- str_trim(x)
  return(x)
}
# Create Short_name column
short_ids <- il12_ids %>%
  # Clean all synonym columns
  mutate(across(c(Name, Synon1, Synon2, Synon3), clean_name)) %>%
  rowwise() %>%
  mutate(
    # Vector of candidate names
    candidate_names = list(
      na.omit(c(Name, Synon1, Synon2, Synon3))),
    # Select shortest valid name
    Short_name = ifelse(
      length(candidate_names) == 0,
      NA_character_,
      candidate_names[which.min(nchar(candidate_names))])) %>%
  ungroup() %>%
  select(-candidate_names)

```

The code below annotated the consensus identification list in the feture list of ions without spectum deconvolution, the code annoted metabolite name comparing the most intensed matched ions with NIST libraries and the retention time with a tolerance of 0.05 min. 

```{r echo=TRUE, message=TRUE, warning=FALSE, error=FALSE}

# Function to parse EI ions
parse_ions <- function(x) {
  if (is.na(x) || x == "") {
    return(numeric(0))}
  as.numeric(str_split(x, ",")[[1]]) %>%
    unique()}
# Main annotation function
add_ids <- function(flist,
                    ids,
                    rt_tol = 0.05) {
  # Prepare identification list
  il_df <- ids %>%
    mutate(ion_list = map(top5_match_mz, parse_ions),
           Annotated_y_n = "No")
  # Prepare GCMS feature list
  gcms_df <- flist %>%
    mutate(row_id_tmp = row_number())
  # Store annotations
  matched_rows <- vector("list", nrow(gcms_df))
  # Main loop
  for (i in seq_len(nrow(gcms_df))) {
    mass_i <- gcms_df$Nominal_mass[i]
    rt_i   <- gcms_df$RT[i]
    col_i  <- gcms_df$Column[i]
    # Find candidates
    candidates <- il_df %>%
      mutate(mass_match = map_lgl(ion_list, ~ any(mass_i %in% .x)),
             rt_match = abs(RT - rt_i) <= rt_tol,
             column_match = Column_type == col_i) %>%
      filter(mass_match,
             rt_match,
             column_match)
    # If matches exist
    if (nrow(candidates) > 0) {
      # Prefer closest RT
      best_match <- candidates %>%
        mutate(rt_diff = abs(RT - rt_i)) %>%
        arrange(rt_diff) %>%
        slice(1)
      # Copy the identification list metadata to the feature list data set row
      matched_rows[[i]] <- bind_cols(gcms_df[i, ],
                                     best_match %>%
                                       select(top5_match_mz,
                                              IL,
                                              Short_name,
                                              Superclass,
                                              Class,
                                              Subclass))
      # Mark metabolite as annotated
      il_df$Annotated_y_n[
        which(il_df$Short_name == best_match$Short_name &
                il_df$RT == best_match$RT)] <- "yes"
    } else {
      matched_rows[[i]] <- gcms_df[i, ]
    }
  }
  # Final annotated GCMS table
  flist_ids <- bind_rows(matched_rows) %>%
    select(-row_id_tmp) %>% 
  relocate(Column, Ion_Mode, .after = last_col())
  # Return results
  return(list(flist_ids = flist_ids,
              flag_ids = il_df %>%
                select(-ion_list)))
}

# Adding the consensus identification list to the feature list
gcms_ms1_fd_ids <- add_ids(flist = gcms_ms1_fd, ids = short_ids, rt_tol = 0.05)

```

In the last code, we added the consensus identification list to the feature list. The result contains two data sets: one with the feature list and the annotated metabolites, and the second with the consensus identification list, including a tag indicating whether each item was annotated. In the following code, we extract the consensus identification list with a flag indicating whether each metabolite was annotated in the feature list. We can inspect whether any metabolite was not identified in R, or we can export the file for further inspection.

```{r echo=TRUE, message=TRUE, warning=FALSE, error=FALSE}

# Extracting the consensus identification list with the annotation flag
tag_ids <- gcms_ms1_fd_ids$flag_ids
# Exporting for the annotation inspection
#write_xlsx(XXXX, "../X_hylaeae_metabolomics/Result/All_column_IDs_flag.xlsx",
#           col_names = FALSE)

```

In the following code, we extract the feature list annotated with the metabolite names. This final annotated feature list will be used to construct the name MetaboSets.

```{r echo=TRUE, message=TRUE, warning=FALSE, error=FALSE}

# Extracting feature list with metabolite annotation
gcms_ms1_fd <- gcms_ms1_fd_ids$flist_ids

```

### Sample information

Some sample information are in the name of the chromatogram files. In the following code we will extract the file name from the exported MZmine feature list, MZmine save the sample information after the 13th column; however we add five columns extra at the bigining of the dataset.

```{r echo=TRUE, message=TRUE, warning=FALSE, error=FALSE}

# Extracting the file names from the MZmine export feature list
file_name <- colnames(gcms_ms1)[18:ncol(gcms_ms1)]
# Extracting the injection order from the file names
inj_order <- sub("_.*", "", file_name)
# Extracting the number of replicates from the file names
sp_repl <- str_match(file_name, "_([^_*]*)\\.mzXML")[, 2] %>%
  str_trim() %>% 
  ifelse(str_detect(., "\\d"), ., NA)
# Extracting the sample group from the file names
sp_gp <- str_match(file_name, "^\\d+_(.*?)_?\\d*\\.mzXML")[, 2] %>%
  str_remove_all("\\s+")
# Sample information from the MZmine feature list
gcms_ms1_pd <- data.frame(Injection_order = inj_order,
                          Group = sp_gp,
                          Visit = sp_repl,
                          Datafile = colnames(gcms_ms1)[18:ncol(gcms_ms1)]) %>%
  # Adding an auxiliary column to keep the original order
  mutate(ax_id = row_number()) %>%
  # Change the Injection_order column value to numeric
  mutate(Injection_order = as.numeric(Injection_order)) %>%
  # Sort globally by group and then by injection (smallest to largest)
  arrange(Group, Injection_order) %>%
  # Group by the "Group" column
  group_by(Group) %>%
  # If 'Visit' is NA, assign the sequence number within the ordered group
  mutate(Visit = ifelse(is.na(Visit), as.character(row_number()), Visit)) %>%
  # Keep the original sequence of the data
  arrange(ax_id) %>%
  select(-ax_id) %>%
  # Ungroup at the end of the process
  ungroup()

```

### Feature abundance

Also the feature abundance of each sample was extracted from the exported MZmine feature list.

```{r echo=TRUE, message=TRUE, warning=FALSE, error=FALSE}

# Extracting feature abundance 
gcms_ms1_xp <- gcms_ms1[, -c(2:17)]
# Change the name of the Global_ID column to ensure the correct merge of data
colnames(gcms_ms1_xp)[colnames(gcms_ms1_xp) == "Global_ID"] <- "Global_ID_xp"

```

### Export file

The notame package uses a defined data format as input; in the following code, we assemble the specific format required by the notame R package.

```{r echo=TRUE, message=TRUE, warning=FALSE, error=FALSE}

# Joining the fdata and xprs
bottom <- cbind(gcms_ms1_fd, gcms_ms1_xp)
# Ensuring that the fdata and xprs data were merged correctly
all(bottom$Global_ID == bottom$Global_ID_xp)
# Once the correct merge of the data is complete, the auxiliary column will be deleted
bottom$Global_ID_xp <- NULL
# Adding the column name
bottom <- bottom %>% dplyr::mutate(across(everything(), 
    as.character)) %>% rbind(colnames(.), .)
# Extracting the column name of the column that has file names
datafile_cols <- colnames(gcms_ms1_pd)[grepl("Datafile", colnames(gcms_ms1_pd))]
# Preparing sample information according to the notame R package format
top <- cbind(matrix(colnames(gcms_ms1_pd), ncol = 1), t(gcms_ms1_pd))
# Adding an empty matrix (size of fData) at the beginning of the pData
empty1 <- matrix(NA_character_, nrow = nrow(top), ncol = ncol(gcms_ms1_fd) - 1)
# Stick an empty matrix (size of fData) with pData
top <- cbind(empty1, top)
# Adding bottom (fData + exprs) column names to the pData
colnames(top) <- colnames(bottom)
# Join the top (pData) and bottom (fData + exprs) datasets
big <- rbind(top[seq_len(nrow(top) - 1), ], bottom)
# Find position of "Injection_order"
start_col <- which(big [1, ] == "Injection_order") + 1
# Convert columns from Injection_order onward as numeric
big[1, start_col:ncol(big)] <- lapply(
  big[1, start_col:ncol(big)],
  function(x) as.numeric(as.character(x)))
# Export file
write_xlsx(big, "../X_hylaeae_metabolomics/Data/notame/GCMS_MS1_apolar_all.xlsx",
           col_names = FALSE)

```

# Notame workflow

Then, a primary path and a log system were added to have a record of each process executed.

```{r echo=TRUE, message=TRUE, warning=FALSE, error=FALSE}

# Main path
ppath <- "../X_hylaeae_metabolomics/"
# Log system
init_log(log_file = paste0(ppath, "Result/notame/GCMS_MS1_log.txt"))

```

Next, the MZmine feature list table in “*notame*” format was loaded.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

data <- read_from_excel("../X_hylaeae_metabolomics/Data/notame/GCMS_MS1_apolar_all.xlsx",
                        corner_row = 4, corner_column = "H",
                        rt_limits = c(0, 57), split_by = c("Column", "Ion_Mode"))

```


Once the data was loaded, the next step was to create a MetaboSet to work with R objects from now on.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

modes <- construct_metabosets(exprs = data$exprs, 
                              pheno_data = data$pheno_data, 
                              feature_data = data$feature_data,
                              group_col = "Group")

```

## Preprocessing

The first step of the preprocessing is to change the features with a value equal to 0 to NA.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Data extraction
mode <- modes$`Rtx-5_EI`
# Change 0 value to NA
mode <- mark_nas(mode, value = 0)





# Boxplot
raw_bp <- plot_sample_boxplots(mode,
                               order_by = "QC",
                               fill_by = "QC")
# PCA
raw_pca <- plot_pca(mode,
                    center = TRUE,
                    shape = "QC",
                    color = "QC")
# Plot
raw_pca + raw_bp

```

Then, features with low detection rates are flagged and can be ignored or removed in subsequent analysis. The “*notame*” package uses two criteria to flag these features: the feature's presence in a percentage of QC injections and the feature's presence in a percentage within a sample group or class.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Low detection rate
mode <- flag_detection(mode, qc_limit = 5/7, group_limit = 3/3)






# Boxplot
det_bp <- plot_sample_boxplots(mode,
                               order_by = "QC",
                               fill_by = "QC")
# PCA
det_pca <- plot_pca(mode,
                    center = TRUE,
                    shape = "QC",
                    color = "QC")
# Plot
det_pca + det_bp


```

The following preprocessing step is drift correction, which is applied using smoothed cubic spline regression.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Drift correction
corrected <- correct_drift(mode)
#corrected <- correct_drift(corrected)
# Flag low quality features
corrected <- flag_quality(corrected, condition = "RSD_r < 0.3 & D_ratio_r < 0.6")

```

Then, we can visualize the data after drift correction.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Boxplot
corr_bp <- plot_sample_boxplots(corrected,
                                order_by = "QC",
                                fill_by = "QC")
# PCA
corr_pca <- plot_pca(corrected,
                     center = TRUE,
                     shape = "QC",
                     color = "QC")
# Package to plots visualization in a same windows
library(patchwork)
# Plot
corr_pca + corr_bp

```

The next step is feature clustering. This step helps us reduce the number of features of the same molecule that were split due to a 70 eV electron ionization (EI) environment.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

clustered <- cluster_features(corrected, rt_window = 1/60, corr_thresh = 0.95,
                              d_thresh = 0.80)
compressed <- compress_clusters(clustered)

```

We can inspect the data using the PCA plot after the clustering algorithm execution.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Boxplot
clust_bp <- plot_sample_boxplots(compressed,
                                 order_by = "QC",
                                 fill_by = "QC")
# PCA
clust_pca <- plot_pca(compressed,
                      center = TRUE,
                      shape = "QC",
                      color = "QC")
# Plot
clust_pca + clust_bp

```

Contaminant peaks based on the process blank will be removed.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# Removal of contaminants
flag_blank <- flag_contaminants(compressed,
                                blank_col = "Group",
                                blank_label = "Ext_blank",
                                flag_thresh = 0.05,
                                flag_label = "Contaminant")
# Removal blank group from dataset
no_blank <- flag_blank[, flag_blank$Group != "Ext_blank"]
pData(no_blank) <- droplevels(pData(no_blank))






# Boxplot
blank_bp <- plot_sample_boxplots(no_blank,
                                 order_by = "QC",
                                 fill_by = "QC")
# PCA
blank_pca <- plot_pca(no_blank,
                      center = TRUE,
                      shape = "QC",
                      color = "QC")
# Plot
blank_pca + blank_bp

```

## Processing

We used probabilistic quotient normalization (PQN) with missing value imputation and scaled using the autoscaling method for downstream statistical analysis.

The code below imputes the data using a chromatographic noise value.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# Impute missing values using noise threshold
imputed <- impute_simple(no_blank, value = 60, na_limit = 0)

```

We can inspect the data with the PCA plot after data imputation.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Boxplot
imp_bp <- plot_sample_boxplots(imputed,
                               order_by = "QC",
                               fill_by = "QC")
# PCA
imp_pca <- plot_pca(imputed,
                    center = TRUE,
                    shape = "QC",
                    color = "QC")
# Plot
imp_pca + imp_bp

```

After data imputation, the data were normalized by PQN.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Probabilistic quotient normalization
pqn_set <- pqn_normalization(imputed,
                             ref = c("qc", "all"),
                             method = c("median", "mean"),
                             all_features = FALSE)

```

We can inspect the data with the PCA plot after data normalization.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Boxplot
pqn_bp <- plot_sample_boxplots(pqn_set,
                               order_by = "QC",
                               fill_by = "QC")
# PCA
pqn_pca <- plot_pca(pqn_set,
                    center = TRUE,
                    shape = "QC",
                    color = "QC")
# Plot
pqn_pca + pqn_bp

```


#################### FALTA EDITAR EN ADELANTE ##################################

Finally, the data is ready to be exported and proceed with the statistical analysis.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

#save(compressed, file = paste0(ppath, "Result/notame_Result/Notame_HS_GC-MS_out.RData"))

```

# Principal component analysis (PCA)

Extracting the data and metadata of "*notame*" MetaboSet for principal component analysis (PCA) plot.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# Extract clean data
pqn_noflag <- drop_flagged(pqn_set)
# Extracting feature height table
peak_height <- exprs(pqn_noflag)
# Extracting phenotypic data
pheno_data <- pqn_noflag@phenoData@data

```

Transposing the feature table and preparing the PCA data.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# Transposing feature height table
transp_table  <- t(peak_height)
# Centering and Scaling features
ei_pca <- prcomp(transp_table, center = TRUE, scale. = TRUE)

```

## Score PCA

Plotting PCA results.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide', fig.width = 7, fig.height = 4}

# Library to left_join use
library(dplyr)
# PCA scores
scores <- ei_pca$x %>%                   # Get PC coordinates
  data.frame %>%                         # Convert to data frames
  mutate(Sample_ID = rownames(.)) %>%    # Create a new column with the sample names
  left_join(pheno_data)                  # Adding metadata
# PCA plot
pca_plot <- ggplot(scores,
       aes(PC1, PC2, shape = Group, color = Group)) +
  scale_color_manual(values=c("#F8766D",
                              "#00BFC4",
                              "#C77CFF",
                              "#7CAE00")) +
  geom_point(size = 3) +
  geom_text(aes(label = DBH_Tag), hjust = 0.5, nudge_x = 0.90, nudge_y = -3.00,
            check_overlap = TRUE, show.legend = FALSE) +
  scale_shape_manual(values=c(16, 15, 3, 17)) +
  guides(x=guide_axis(title = "PC1 (23.83 %)"),
         y=guide_axis(title = "PC2 (18.50 %)")) +
  labs(shape = 'DBH', color= 'DBH') +
  theme_classic() +
  theme(legend.text = element_text(face = "plain")) +
  theme(legend.position = c(0.120, 0.760),
        legend.background = element_rect(fill = "white", color = "black")) +
  theme(panel.grid = element_blank(), 
        panel.border = element_rect(fill= "transparent")) +
  geom_vline(xintercept = 0, linetype = "longdash", colour="gray") +
  geom_hline(yintercept = 0, linetype = "longdash", colour="gray")
pca_plot

```

## Loading PCA

Plotting a loading PCA result.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

loadings <- ei_pca$rotation %>%           # Extract loadings
  data.frame(Feature_ID = rownames(.))    # New column with feat name

```

Creating an artificial table with feature names and a column with compound names (identified metabolites).

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide', fig.width = 14, fig.height = 7}

# Extracting feature identified at level 2
metab_data <- pqn_noflag[!is.na(pqn_noflag@featureData@data$Identification_level),]
metab_data <- metab_data[metab_data@featureData@data$Identification_level == 
                           "Identification level 2",]
# Extracting metabolite table
meta_table <- metab_data@featureData@data
# Creating a new small table of the annotated compounds
ei_compouds <- left_join(meta_table, loadings)
# Plotting results
load_pca <- ggplot(loadings, aes(PC1, PC2)) + 
  geom_point(alpha = 0.3, size = 2.5) +
  theme_classic() +
ggrepel::geom_label_repel(data = ei_compouds,
                            aes(label = meta_table$Metabolite),
                            box.padding = 0.37,
                            label.padding = 0.22,
                            label.r = 0.30,
                            cex = 3,
                            max.overlaps = 50,
                            min.segment.length = 0) +
  guides(x=guide_axis(title = "PC1 (23.83 %)"),
         y=guide_axis(title = "PC2 (18.50 %)")) +
  #theme(legend.position = c(0.070, 0.090),
  #      legend.background = element_rect(fill = "white", color = "black")) +
  theme(panel.grid = element_blank(), 
        panel.border = element_rect(fill= "transparent")) +
  geom_vline(xintercept = 0, linetype = "longdash", colour="gray") +
  geom_hline(yintercept = 0, linetype = "longdash", colour="gray")
load_pca
# Save plot
# Save plot in *.pdf
#ggsave(filename = '../Headspace_of_C_lechleri_sap/Result/notame_Result/loading_pca_1A.pdf',
#       plot = load_pca, width = 14, height = 7, dpi = 300, scale = 0.85)
# Save plot in *.png
#ggsave(filename = '../Headspace_of_C_lechleri_sap/Result/notame_Result/loading_pca_1A.png',
#       plot = load_pca, width = 14, height = 7, dpi = 300, scale = 0.85)
# Save plot in *.jpg
#ggsave(filename = '../Headspace_of_C_lechleri_sap/Result/notame_Result/loading_pca_1A.jpg',
#       plot = load_pca, width = 14, height = 7, dpi = 300, scale = 0.85)

```

# Heatmap with HCA

Only the annotated features were used in the heatmap and the hierarchical cluster analysis (HCA). The ComplexHeatmap package will be used for the heatmap and the hierarchical cluster analysis (HCA).

ComplexHeatmap package and dependency installation.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# ComplexHeatmap package installation
#if (!requireNamespace("BiocManager", quietly=TRUE))
#    install.packages("BiocManager")
#BiocManager::install("ComplexHeatmap")
library(ComplexHeatmap)

# ColorRamp2 package installation
#if (!requireNamespace("devtools", quietly = TRUE)) {
#  install.packages("devtools")
#}
#devtools::install_github("jokergoo/colorRamp2")
library(colorRamp2)

# Cowplot package installation
#install.packages("cowplot")
library(cowplot)

# mdatools package installation
#install_github('svkucheryavski/mdatools')
library(mdatools)

# ClassyFire package installation
#remotes::install_github('aberHRML/classyfireR')
library(classyfireR)

```

The metabolites were classified using the [ClassyFireR](https://doi.org/10.1186/s13321-016-0174-y) to add these metabolite classifications to the heatmap plot.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# InChI key of the metabolites you want to classify
InChI_Keys <- c('2-Heptanone' = "KYWIYKKSMDLRDC-UHFFFAOYSA-N")
# Get classification
Classification_List <- purrr::map(InChI_Keys, get_classification)
Classification_List

``` 

Extracting and loading of identified metabolites' abundance, and data scaling by the autoscaling method.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Drop QC
hm_no_qc <- drop_qcs(pqn_noflag)
# Scaling by autoscaling method
hm_scl <- scale(t(exprs(hm_no_qc)), center = TRUE, scale = TRUE)
hm_scl <- t(hm_scl)
# Adding autoscaling data to notame MetaboSet
hm_scl_set <- hm_no_qc
exprs(hm_scl_set) <- hm_scl
# Extracting identified metabolite data at level 2
raw_hm <- hm_scl_set[!is.na(hm_scl_set@featureData@data$Metabolite),]
raw_hm <- raw_hm[raw_hm@featureData@data$Identification_level == 
                   "Identification level 2",]
# Extracting feature height table
hm_height <- exprs(raw_hm)
# Extracting sample information
hm_pdata <- raw_hm@phenoData@data
# Extracting feature information
hm_fdata <- raw_hm@featureData@data

```

Row and top heatmap annotation.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, fig.width = 10, fig.height = 9}

set.seed(1540)
# Adding row and column names
hm_scl <- hm_height
rownames(hm_scl) <- hm_fdata$Metabolite
colnames(hm_scl) <- hm_pdata$DBH_Tag
# Metabolite class color
cols_metclass <- c("Monoterpenoids" = "#800000FF",
                   "Fatty alcohols" = "#8DD3C7",
                   "Miscellaneous" = "#C0C0C0",
                   "Sesquiterpenoids" = "#FFA319FF",
                   "Carbonyl compounds" = "#EED5B7",
                   "Alkanes" = "#BEBADA")
# Add row anotation to HeatMap
hm_row_ann <- rowAnnotation(Subclass = hm_fdata$classyfireR_Subclass,
                            col = list(Subclass = cols_metclass),
                            show_annotation_name = TRUE,
                            show_legend = FALSE)
# Species color
cols_age <- c("92-96 cm" = "#E76BF3",
              "108-112 cm" = "#F8766D",
              "79-86 cm" = "#7CAE00")
# Add top anotation to Heatmap
top_info_ann <- HeatmapAnnotation(`Plant age` = hm_pdata$Group,
                                  col = list(`Plant age` = cols_age),
                                  show_annotation_name = T,
                                  show_legend = F,
                                  border = T)
# Color scale
mycol <- colorRamp2(c(-4, 0, 4), c("blue", "white", "red"))
# Heatmap matrix plotting
hm_plot <- Heatmap(hm_scl,
                   col = mycol,
                   border_gp = grid::gpar(col = "black", lty = 0.02),
                   rect_gp = grid::gpar(col = "black", lwd = 0.75),
                   clustering_distance_columns = "euclidean",
                   clustering_method_columns = "complete",
                   top_annotation = top_info_ann,
                   row_names_max_width = unit(10, "cm"),
                   right_annotation = hm_row_ann,
                   show_heatmap_legend = F,
                   row_km = 3, column_km = 3,
                   row_title = c("a", "b", "c"))

hm_plot

```

Adding legends to the heatmap.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Color scale legend
lgd1 <- Legend(col_fun = mycol,
               title = "Autoscaled abundance",
               direction = "horizontal" )
# Bees species legend
lgd2 <- Legend(labels = gt_render(c("92-96 cm",
                                    "108-112 cm",
                                    "79-86 cm")),
               legend_gp = gpar(fill = cols_age),
               title = "DBH", ncol = 1)
# Metabolite class Legend
lgd3 <- Legend(labels = c(unique(hm_fdata$classyfireR_Subclass)) ,
               legend_gp = gpar(fill = cols_metclass), 
               title = "Metabolite Subclass", ncol = 2)

```

Plotting the heatmap.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, fig.width = 10, fig.height = 12}

set.seed(1540)
# Converting to ggplot
gg_heatmap <- grid.grabExpr(draw(hm_plot))
gg_heatmap <- ggpubr::as_ggplot(gg_heatmap)
# Legends
all_legends <- packLegend(lgd1, lgd2, lgd3, direction = "horizontal")
gg_legend <- grid.grabExpr(draw(all_legends))
gg_legend_fn <- ggpubr::as_ggplot(gg_legend)
# Heatmap plot
gcms_hm <- plot_grid(gg_legend_fn,
                     gg_heatmap, ncol = 1,
                     rel_heights = c(0.055, 0.880))
gcms_hm

```

# Antioxidant activity correlation

Correlation between feature abundance and antioxidant result (TAC, total antioxidant content) using the PCA and Pearson's correlation approaches.

## Using the PCA

Preparation of data and metadata, merging antioxidant activity data with abundance of features.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE} 

# Antioxidant activity data extraction
Antioxidant_data <- drop_qcs(modes$na_na)
Antioxidant_data <- t(exprs(Antioxidant_data))
# Drop QCs of metabolomics data
no_qc <- drop_qcs(pqn_noflag)
# Extracting feature height table
corr_peak <- t(exprs(no_qc))
# Merge the antioxidant activity result with the feature list table
Antioxidant_peak <- cbind(Antioxidant_data, corr_peak)
# Extracting phenotypic data of metabolomics data
corr_pdata <- no_qc@phenoData@data

```

### Score PCA

PCA scores calculation of the antioxidant and feature abundance merged data.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Centering and Scaling features
corr_pca <- prcomp(Antioxidant_peak, center = TRUE, scale. = TRUE)

```

Plotting PCA score.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide', fig.width = 7, fig.height = 4}

# PCA scores
corr_scores <- corr_pca$x %>%            # Get PC coordinates
  data.frame %>%                         # Convert to data frames
  mutate(Sample_ID = rownames(.)) %>%    # Create a new column with the sample names
  left_join(corr_pdata)                  # Adding metadata
# PCA plot
corr_pca_plot <- ggplot(corr_scores,
                        aes(PC1, PC2, shape = Group, color = Group)) +
  scale_color_manual(values=c("#F8766D",
                              "#00BFC4",
                              "#C77CFF")) +
  scale_shape_manual(values=c(16, 15, 3)) +
  geom_point(size = 3) +
  geom_text(aes(label = DBH_Tag), hjust = 0.5, nudge_x = 0.90, nudge_y = -3.00,
            check_overlap = TRUE, show.legend = FALSE) +
  guides(x=guide_axis(title = "PC1 (27.01 %)"),
         y=guide_axis(title = "PC2 (21.27 %)")) +
  labs(shape = 'DBH', color= 'DBH') +
  theme_classic() +
  #theme(legend.text = element_text(face="italic")) +
  theme(legend.position = c(0.120, 0.795),
        legend.background = element_rect(fill = "white", color = "black")) +
  theme(panel.grid = element_blank(), 
        panel.border = element_rect(fill= "transparent")) +
  geom_vline(xintercept = 0, linetype = "longdash", colour="gray") +
  geom_hline(yintercept = 0, linetype = "longdash", colour="gray")
corr_pca_plot

```

### Loading PCA

Plotting loading results.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

corr_loadings <- corr_pca$rotation %>%    # Extract loadings
  data.frame(Feature_ID = rownames(.))    # New column with feat name

```

Creating an artificial table with feature names and a column with compound names (identified metabolites).

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide', fig.width = 14, fig.height = 7}

# Extracting the Antioxidant metadata
Antioxidant_table <- drop_qcs(modes$na_na)@featureData@data
Antioxidant_table$Flag  <- as.character(Antioxidant_table$Flag)
# Merge the Antioxidant metadata with the metabolite table
corr_table <- merge(meta_table, Antioxidant_table, all = TRUE)
# Creating a new small table of the annotated compounds and antimicrobial metadata
corr_compouds <- left_join(corr_table, corr_loadings)
# Plotting results
corr_load_plot <- ggplot(corr_loadings, aes(PC1, PC2)) + 
  geom_point(alpha = 0.3, size = 2) +
  theme_classic() +
  geom_point(data = corr_compouds, aes(shape = corr_table$Identification_level,
                                       color = corr_table$Identification_level), size = 2) +
  labs(shape = 'Legend of features',
       color = 'Legend of features') +
  scale_color_manual(values = c("red",
                                "darkblue")) +
  scale_shape_manual(values = c(17, 19)) +
  ggrepel::geom_label_repel(data = corr_compouds,
                            aes(label = corr_table$Metabolite_name),
                            box.padding = 0.37,
                            label.padding = 0.22,
                            label.r = 0.30,
                            cex = 2.5,
                            max.overlaps = 50,
                            min.segment.length = 0) +
  guides(x=guide_axis(title = "PC1 (27.01 %)"),
         y=guide_axis(title = "PC2 (21.27 %)")) +
  theme(legend.position = c(0.075, 0.890),
        legend.background = element_rect(fill = "white", color = "black")) +
  theme(panel.grid = element_blank(), 
        panel.border = element_rect(fill= "transparent")) +
  geom_vline(xintercept = 0, linetype = "longdash", colour="gray") +
  geom_hline(yintercept = 0, linetype = "longdash", colour="gray")
  #ggsci::scale_color_aaas()
corr_load_plot

```

## Using the Pearson correlation

The Pearson correlation between Antioxidant activity and feature abundance was performed.

### Normality of the Antioxidant data

First, the Antioxidant data normality was evaluated using the Shapiro-Wilk test.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Loading the "readxl" package
library(readxl)
# Loading the Antioxidant activity data
all_Antioxidant <- read_excel("../Headspace_of_C_lechleri_sap/Data/C_lechleri_sap_Antioxidant.xlsx",
                              sheet = 2)
# Deleting the group column to perform Shapiro-Wilk test in batch
no_group <- all_Antioxidant[,-1]
# Perform the Shapiro-Wilk test
shapiro <- do.call(rbind, lapply(no_group,
                                 function(x) shapiro.test(x)[c("statistic", "p.value")]))
shapiro

```

The result showed that only the antioxidant activities ABTS and DPPH have a normal distribution (p-value > 0.05).

### Metabolomics data transformation

Before the Pearson correlation of feature abundance with the Antioxidant activity (Total antioxidant content), the previously normalized data by PQN was transformed using the generalised logarithm (glog) method.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# "SummarizedExperiment" package installation
#if (!require("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")
#BiocManager::install("SummarizedExperiment")
library(SummarizedExperiment)
# Convert feature height table to SummarizedExperiment class
pmp_data <- SummarizedExperiment(assays = exprs(pqn_noflag),
                                 colData = pqn_noflag@phenoData@data)
# Package for generalized logarithmic transform
#if (!requireNamespace("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")
#BiocManager::install("pmp")
library(pmp)
# Generalised logarithmic transform
glog_exprs <- glog_transformation(df = pmp_data@assays@data@listData[[1]],
                                  classes = pmp_data$QC,
                                  qc_label = "QC")
# Adding glog transformation to notame MetaboSet
glog_set <- pqn_noflag
exprs(glog_set) <- glog_exprs

```

Preparing the metabolomics data for the Antioxidant correlation.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Drop QCs
glog_no_qc <- drop_qcs(glog_set)
# Convert Antioxidant result to numeric
glog_no_qc$ABTS <- as.numeric(glog_no_qc$ABTS)
glog_no_qc$DPPH <- as.numeric(glog_no_qc$DPPH)
glog_no_qc$DBH_split10 <- as.numeric(glog_no_qc$DBH_split10)

```

Perform the Pearson correlation analysis.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Feature correlation with ABTS antioxidant activity
ABTS_corr <- perform_correlation_tests(glog_no_qc, x = featureNames(glog_no_qc),
                                       y = c("ABTS"), method = "pearson")

# Feature correlation with DPPH antioxidant activity
DPPH_corr <- perform_correlation_tests(glog_no_qc, x = featureNames(glog_no_qc),
                                       y = c("DPPH"), method = "pearson")

# Feature correlation with DBH
DBH_corr <- perform_correlation_tests(glog_no_qc, x = featureNames(glog_no_qc),
                                       y = c("DBH_split10"), method = "pearson")

```

### Correlation of identified metabolites

Create a matrix for the heatplot of Pearson's correlation.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# Correlation table for the heatmap
hm_corr_table <- data.frame(Feature_ID = ABTS_corr$X,
                            ABTS = ABTS_corr$Correlation_coefficient,
                            DPPH = DPPH_corr$Correlation_coefficient,
                            DBH = DBH_corr$Correlation_coefficient)
# Feature data table
feat_corr_table <- data.frame(Feature_ID = hm_fdata$Feature_ID,
                             Metabolite = hm_fdata$Metabolite)
# Adding the metabolie name to the correlation table
hm_corr_table <- left_join(hm_corr_table, feat_corr_table)
# Extracting data of the identified metabolites
hm_corr_table <- hm_corr_table[!is.na(hm_corr_table$Metabolite),]
# Adding row name
rownames(hm_corr_table) <- hm_corr_table$Metabolite
# Delete extra information
hm_corr_table <- subset(hm_corr_table, select = -c(Feature_ID, Metabolite))
# Converting DataFrame to data matrix
hm_corr_table <- data.matrix(hm_corr_table, rownames.force = NA)
# Adding column name
colnames(hm_corr_table) <- c("ABTS", "DPPH", "DBH")

```

Create a matrix showing the statistical significance of the correlation result.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# Correlation table for the heatmap
hm_sign_table <- data.frame(Feature_ID = ABTS_corr$X,
                            ABTS = ABTS_corr$Correlation_P,
                            DPPH = DPPH_corr$Correlation_P,
                            DBH = DBH_corr$Correlation_P)
# Adding the metabolie name to the significance table
hm_sign_table <- left_join(hm_sign_table, feat_corr_table)
# Extracting data of the identified metabolites
hm_sign_table <- hm_sign_table[!is.na(hm_sign_table$Metabolite),]
# Adding row name
rownames(hm_sign_table) <- hm_sign_table$Metabolite
# Delete extra information
hm_sign_table <- subset(hm_sign_table, select = -c(Feature_ID, Metabolite))
# Converting DataFrame to data matrix
hm_sign_table <- data.matrix(hm_sign_table, rownames.force = NA)
# Adding column name
colnames(hm_sign_table) <- c("ABTS", "DPPH", "DBH")

```

Plot the correlation heatplot.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, fig.width = 5, fig.height = 9}

# Add top anotation to Heatmap
ann_corr <- HeatmapAnnotation(
  foo = anno_block(gp = gpar(fill = 0, col = "white"),
                   labels = "Pearson\n correlation",
                   labels_gp = gpar(col = "black", fontsize = 12, fontface = "bold")))
# Color scale
col_fun <- colorRamp2(c(-1, 0, 1), c("blue", "white", "red"))
# Heatmap plot
hm_corr <- Heatmap(hm_corr_table, col = col_fun, cluster_rows = FALSE,
                   cluster_columns = FALSE,
                   right_annotation = hm_row_ann,
                   border_gp = grid::gpar(col = "black", lty = 0.02),
                   rect_gp = grid::gpar(col = "black", lwd = 0.75),
                   top_annotation = ann_corr,
                   cell_fun = function(j, i, x, y, w, h, fill) {
                     if(hm_sign_table[i, j] < 0.05) {
                       grid.text(sprintf("%.1f", hm_corr_table[i, j]), x, y,
                                 gp = gpar(fontsize = 9, fontface = "bold"))
                       } else if(hm_sign_table[i, j] > 0.05) {
                         grid.text(sprintf("%.1f", hm_corr_table[i, j]), x, y,
                                   gp = gpar(fontsize = 9))}},
                   show_heatmap_legend = F)
hm_corr

```

### Correlation of unknown metabolites

Filtering the metabolites with positive and high Pearson's correlation coefficient (r >0.5).

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Library to change character values
library('stringr')
# Correlation table for the heatmap with the best features
best_feat <- data.frame(Feature_ID = ABTS_corr$X,
                        ABTS = ABTS_corr$Correlation_coefficient,
                        ABTS_p = ABTS_corr$Correlation_P,
                        DPPH = DPPH_corr$Correlation_coefficient,
                        DPPH_p = DPPH_corr$Correlation_P)
# Filtering the feature with the correlation >= 0.7
best_feat <- best_feat[abs(best_feat$ABTS) >= 0.7 |
                         abs(best_feat$DPPH) >= 0.7, ]
# Adding the metabolite name to the correlation table
best_feat <- left_join(best_feat, feat_corr_table)
# Extracting data of no identified metabolites
best_feat <- best_feat[!complete.cases(best_feat$Metabolite),]
# Replace the column name and ionization type "RTX5MS_EI" with unknown
best_feat$Feature_ID <- sub('SH-RTX-5Sil-MS_EI_','Unknown-', best_feat$Feature_ID)
best_feat$Feature_ID <- str_replace_all(best_feat$Feature_ID, '_', '.')

```

Create a matrix for the heatplot of Pearson's correlation.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# Correlation table for the heatmap
hm_corr_table1 <- data.frame(Feature_ID = best_feat$Feature_ID,
                             ABTS = best_feat$ABTS,
                             DPPH = best_feat$DPPH )
# Adding row name
rownames(hm_corr_table1) <- hm_corr_table1$Feature_ID
# Delete extra information
hm_corr_table1 <- subset(hm_corr_table1, select = -c(Feature_ID))
# Converting DataFrame to data matrix
hm_corr_table1 <- data.matrix(hm_corr_table1, rownames.force = NA)
# Adding column name
colnames(hm_corr_table1) <- c("ABTS", "DPPH")

```

Create a matrix showing the statistical significance of the correlation result.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# Correlation table for the heatmap
hm_sign_table1 <- data.frame(Feature_ID = best_feat$Feature_ID,
                             ABTS = best_feat$ABTS_p,
                             DPPH = best_feat$DPPH_p)
# Adding row name
rownames(hm_sign_table1) <- hm_sign_table1$Feature_ID
# Delete extra information
hm_sign_table1 <- subset(hm_sign_table1, select = -c(Feature_ID))
# Converting DataFrame to data matrix
hm_sign_table1 <- data.matrix(hm_sign_table1, rownames.force = NA)
# Adding column name
colnames(hm_sign_table1) <- c("ABTS", "DPPH")

```

Plot the correlation heatplot.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, fig.width = 5, fig.height = 6}

# Color scale legend
lgd1a <- Legend(col_fun = col_fun,
                title = "Correlation coefficient",
                direction = "horizontal" )
# Packed legends
pd = packLegend(lgd1a, direction = "horizontal")
# Heatmap plot
#png(file="Result/notame_Result/HS_GCMS/FigureA1.png", width = 5, height = 6, units = "in", res= 300)
hm_corr1 <- Heatmap(hm_corr_table1, col = col_fun, cluster_rows = FALSE,
                    cluster_columns = FALSE,
                    border_gp = grid::gpar(col = "black", lty = 0.02),
                    rect_gp = grid::gpar(col = "black", lwd = 0.75),
                    top_annotation = ann_corr,
                    cell_fun = function(j, i, x, y, w, h, fill) {
                     if(hm_sign_table1[i, j] < 0.05) {
                       grid.text(sprintf("%.1f", hm_corr_table1[i, j]), x, y,
                                 gp = gpar(fontsize = 10, fontface = "bold"))
                       } else if(hm_sign_table1[i, j] > 0.05) {
                         grid.text(sprintf("%.1f", hm_corr_table1[i, j]), x, y,
                                   gp = gpar(fontsize = 10))}},
                    show_heatmap_legend = F)
hm_corr1
draw(pd, x = unit(7.5, "cm"), y = unit(0.2, "cm"), just = c("left", "bottom"))
#dev.off()

```

Finish a record.

```{r}

finish_log()

```
