---
title: "GC-MS metabolomics of *in vitro* *X. hylaeae* metabolomics- Spectral deconvolution"
author: "Jefferson Pastuna"
date: "2026-04-15"
output:
  github_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
usethis::git_vaccinate()

```

### Introduction

Write an Introduction.

### Before to start

Write a package description.

### eRah package workflow

How to star.

```{r echo=TRUE, warning=FALSE}

# eRah package installation
#install.packages('erah')
# eRah library call
library(erah)

```

Folder files

```{r echo=TRUE}

# Delete all file that are not in folders
unlink('Data/eRah_polar_colum/*')
# Data folder path
createdt('Data/eRah_polar_colum/')

```

New experiment.


```{r echo=TRUE}

# Loading (*.CDF) chromatograms
instrumental <- read.csv('Data/eRah_metadata_polar_column/eRah_polar_colum_inst.csv', sep = ";")
# Loading metadata of the chromatograms (phenotype file)
phenotype <- read.csv('Data/eRah_metadata_polar_column/eRah_polar_colum_pheno.csv', sep = ";")
# Merge of metadata information with chromatograms
ex <- newExp(instrumental = instrumental,
             phenotype = phenotype,
             info = 'X. hylaeae metabolomics Wax colmun')

```

Compound deconvolution

Parameter

```{r echo=TRUE}

# Deconvolution parameters for chromatogram analysis between 3.7 and 38.0 min
dec.par.3to38 <- setDecPar(min.peak.width = 3,
                           min.peak.height = 50,
                           noise.threshold = 500,
                           avoid.processing.mz = c(30:32),
                           analysis.time = c(3.7, 38.0))
# Deconvolution parameters for chromatogram analysis between 38.0 and 50.9 min
dec.par.38to50 <- setDecPar(min.peak.width = 3,
                           min.peak.height = 800,
                           noise.threshold = 80,
                           avoid.processing.mz = c(30:32),
                           analysis.time = c(38.0, 50.9))

```

parallel processing

```{r echo=TRUE}

plan(future::multisession, workers = 5)

```

Deconvolution

```{r echo=TRUE, warning=FALSE}

# Chromatogram deconvolution between 3.7 and 38.0 min
ex.3to38 <- deconvolveComp(ex, dec.par.3to38)
# Chromatogram deconvolution between 38.0 and 50.9 min
ex.38to50 <- deconvolveComp(ex, dec.par.38to50)

```

Alignment

```{r echo=TRUE}

# Alignment parameters
ex.al.par <- setAlPar(min.spectra.cor = 0.85,
                      max.time.dist = 3,
                      mz.range = 50:500)
# Alignment of 3.7-38.0 min dataset
ex.3to38 <- alignComp(ex.3to38, alParameters = ex.al.par)
# Alignment of 38.0-50.9 min dataset
ex.38to50 <- alignComp(ex.38to50, alParameters = ex.al.par)

```

Missing compound recovery

```{r echo=TRUE}

# Missing compound recovery for 3.7-38.0 min dataset
ex.3to38 <- recMissComp(ex.3to38, min.samples = 2)
# Missing compound recovery for 38.0-50.9 min dataset
ex.38to50 <- recMissComp(ex.38to50, min.samples = 2)

```

Exporting alignment feature list.

```{r echo=TRUE}

# Extracting alignment feature list 3.7-38.0 min
feat_list.3to38 <- ex.3to38@Results@Alignment
# Exporting alignment feature list
write.csv(feat_list.3to38,
          file = "Result/eRah_3to38min_GCMS_polar/eRah_feat_list_3to38min.csv")
# Extracting alignment feature list 38.0-50.9 min
feat_list.38to50 <- ex.38to50@Results@Alignment
# Exporting alignment feature list
write.csv(feat_list.38to50,
          file = "Result/eRah_38to50min_GCMS_polar/eRah_feat_list_38to50min.csv")

```

# Identification

Metabolite identification was by comparing all the spectra found against a reference database. The eRah default database was used to metabolite identification. Metabolite identification was improve by exporting all the spectra found (.msp) in eRah to NIST MS Search 2.4 software.

```{r echo=TRUE, warning=FALSE}

# Identification of 3.7-38.0 min dataset
peak_iden.3to38 <- identifyComp(ex.3to38,
                                id.database = mslib,
                                mz.range = NULL,
                                n.putative = 1)
# Identification of 38.0-50.9 min dataset
peak_iden.38to50 <- identifyComp(ex.38to50,
                                 id.database = mslib,
                                 mz.range = NULL,
                                 n.putative = 1)

```

Exporting spectra to NIST MS Search software identification with NIST-20 library.

```{r echo=TRUE}

# 3.7-38.0 min dataset
export2MSP(peak_iden.3to38,
           store.path = "Result/eRah_3to38min_GCMS_polar",
           alg.version = 2)
# 38.0-50.9 min dataset
export2MSP(peak_iden.38to50,
           store.path = "Result/eRah_38to50min_GCMS_polar",
           alg.version = 2)

```

