---
title: "Procedures to export NIST 20 libraries (*mainlib* and *replib*) using NIS MS Search and them loading the libraries into R"
author: "Jefferson Pastuna"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
usethis::git_vaccinate()

```

## Libraries exportation from NIST MS Search

The *mainlib* and *replib* libraries was exported from NIST MS Search following the procedures described in https://github.com/QizhiSu/mspcompiler. Libraries was exported in (*.TXT) format, this file will be used to load in R.

## Loading NIST libraries in R

Loading the necesary libraries.

```{r }

library(stringr)
library(dplyr)
library(tidyr)
library(readxl)
library(writexl)
library(purrr)

```

The NIST data base (*mainlib* and *replib*) was load to extract metadata like, metabolite name, synomins, InChlkey, CAS, and RI.

```{r }

# Function to load NIST libraries *.txt
parse_nist_txt <- function(file_path) {
  lines <- readLines(file_path, warn = FALSE)
  # Split by compounds
  blocks <- split(lines, cumsum(grepl("^Name:", lines)))
  parse_block <- function(block) {
    block_txt <- paste(block, collapse = "\n")
    # Principal fields
    name <- str_match(block_txt, "Name:\\s*(.*)")[,2]
    inchikey <- str_match(block_txt, "InChIKey:\\s*(.*)")[,2]
    # Synon (only 3)
    synon_all <- str_match_all(block_txt, "Synon:\\s*(.*)")[[1]][,2]
    synon1 <- if(length(synon_all) >= 1) synon_all[1] else NA
    synon2 <- if(length(synon_all) >= 2) synon_all[2] else NA
    synon3 <- if(length(synon_all) >= 3) synon_all[3] else NA
    # CAS y Related CAS
    cas <- str_match(block_txt, "CAS#:\\s*([^;\\n]+)")[,2]
    related_cas <- str_match(block_txt, "Related_CAS#:\\s*([^;\\n]+)")[,2]
    # NIST ID
    nist_id <- str_match(block_txt, "NIST#:\\s*(\\d+)")[,2]
    # DB number
    db_id <- str_match(block_txt, "DB#:\\s*(\\d+)")[,2]
    # Retention Index
    ri_line <- str_match(block_txt, "Retention_index:\\s*(.*)")[,2]
    ri_semi <- NA
    ri_stdnp <- NA
    ri_polar <- NA
    if(!is.na(ri_line)) {
      parts <- str_split(ri_line, "\\s+")[[1]]
      for(p in parts) {
        # SemiStdNP
        if(str_detect(p, "SemiStdNP=")) {
          val <- str_match(p, "SemiStdNP=(\\d+)")[,2]
          if(is.na(val)) val <- str_match(p, "SemiStdNP=(\\d+)/")[,2]
          ri_semi <- as.numeric(val)}
        # StdNP
        if(str_detect(p, "StdNP=")) {
          val <- str_match(p, "StdNP=(\\d+)")[,2]
          if(is.na(val)) val <- str_match(p, "StdNP=(\\d+)/")[,2]
          ri_stdnp <- as.numeric(val)}
        # StdPolar
        if(str_detect(p, "StdPolar=")) {
          val <- str_match(p, "StdPolar=(\\d+)")[,2]
          if(is.na(val)) val <- str_match(p, "StdPolar=(\\d+)/")[,2]
          ri_polar <- as.numeric(val)
        }
      }
    }
    tibble(
      Name = name,
      Synon1 = synon1,
      Synon2 = synon2,
      Synon3 = synon3,
      InChIKey = inchikey,
      CAS = cas,
      Related_CAS = related_cas,
      NIST_ID = nist_id,
      DB_ID = db_id,
      RI_SemiStdNP = ri_semi,
      RI_StdNP = ri_stdnp,
      RI_StdPolar = ri_polar
    )
  }
  map_dfr(blocks, parse_block)
}
# Cargar bibliotecas
mainlib <- parse_nist_txt("F:/NIST_20_Library/Data/mainlib.txt")
replib  <- parse_nist_txt("F:/NIST_20_Library/Data/replib.txt")
ikiamlib  <- parse_nist_txt("F:/NIST_20_Library/Data/ikiamlib.txt")
# Unir
nist_db <- bind_rows(
  mainlib %>% mutate(Library = "mainlib"),
  replib  %>% mutate(Library = "replib"),
  ikiamlib %>% mutate(Library = "ikiam_npl_gcms_lib"))
# Ensuring a unic compound in the data base
nist_clean <- nist_db %>%
  distinct(DB_ID, Library, .keep_all = TRUE)
# Saving the processed database
saveRDS(nist_clean, "../X_hylaeae_metabolomics/Result/nist_db.rds")

```

In order to protect the licenced libraries policy, we only provide the code used to process the exported NIST 20 .TXT libraries (*mainlib* and *replib*) to manage in R. This data will be used add the metabolite class in the identified metabolites using NIST MS Search and experimentar algoritm deconvoluted EI spectrum.

The following code extract the EI mass espectrum from NIST 20 libraries (*mainlib* and *replib*). The EI mass espectrum will be used to count the number of ions match with the experimental deconvoluted softwares like MS-DIAL, MZmine, eRah and MSHub. Count the number of ions between libraries and experiment espectrum support the identification match factor reported by the NIST MS Search program.

```{r }

# Function to load NIST libraries *.txt
parse_nist_mz <- function(file_path) {
  lines <- readLines(file_path, warn = FALSE)
  blocks <- split(lines, cumsum(grepl("^Name:", lines)))
  blocks <- blocks[lengths(blocks) > 1]
  parse_block <- function(block) {
    txt <- paste(block, collapse = "\n")
    # DB_ID
    db_id <- str_match(txt, "DB#:\\s*(\\d+)")[,2]
    # Numbers of Peaks
    num_peaks <- str_match(txt, "Num Peaks:\\s*(\\d+)")[,2]
    # Extrac spectrum
    peaks_lines <- block[grepl("^\\d+\\s", block)]
    if(length(peaks_lines) == 0) return(NULL)
    peaks_txt <- paste(peaks_lines, collapse = " ")
    # Split mass m/z and intensity
    pairs <- unlist(str_split(peaks_txt, ";"))
    pairs <- trimws(pairs)
    pairs <- pairs[pairs != ""]
    mz <- as.numeric(str_extract(pairs, "^\\d+"))
    intensity <- as.numeric(str_extract(pairs, "\\d+$"))
    tibble(
      DB_ID = as.character(db_id),
      nist_mz = list(mz),
      nist_intensity = list(intensity)
    )
  }
  map_dfr(blocks, parse_block)
}
# Loading NIST libraries
mainlib_mz <- parse_nist_mz("F:/NIST_20_Library/Data/mainlib.txt")
replib_mz  <- parse_nist_mz("F:/NIST_20_Library/Data/replib.txt")
ikiamlib_mz  <- parse_nist_mz("F:/NIST_20_Library/Data/ikiamlib.txt")
# Merge all libraries
nist_mz <- bind_rows(
  mainlib_mz %>% mutate(Library = "mainlib"),
  replib_mz  %>% mutate(Library = "replib"),
  ikiamlib_mz %>% mutate(Library = "ikiam_npl_gcms_lib"))
# Ensuring a unic compound in the data base
nist_mz <- nist_mz %>%
  distinct(DB_ID, Library, .keep_all = TRUE)
# Saving the processed database to easy loading in the R
saveRDS(nist_mz, "../X_hylaeae_metabolomics/Result/nist_mz.rds")

```

In order to protect the licenced libraries policy of NIST, we only provide the code used to process the exported NIST 20 .TXT libraries (*mainlib* and *replib*) to manage in R.
