---
title: "Metabolite identified using NIST MS Search"
author: "Jefferson Pastuna"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

```

## Introduction

This notebook aim show how the hits result from NIST MS Search using NIST 20 library was filtered in order tu report metabolites with the best match and small RI shift.

### MS-DIAL

Loading the necesary libraries.

```{r }

library(stringr)
library(dplyr)
library(tidyr)
library(readxl)
library(writexl)
library(purrr)

```

Reading the files.

```{r }

file_md3t49m <- "../X_hylaeae_metabolomics/Data/NIST_MS_Search/MSDIAL_3-49min.TXT"
file_md49t54m <- "../X_hylaeae_metabolomics/Data/NIST_MS_Search/MSDIAL_49-54min.TXT"
file_md54t56m <- "../X_hylaeae_metabolomics/Data/NIST_MS_Search/MSDIAL_54-56min.TXT"
md3t49m <- readLines(file_md3t49m, encoding = "latin1")
md49t54m <- readLines(file_md49t54m, encoding = "latin1")
md54t56m <- readLines(file_md54t56m, encoding = "latin1")

```

Formating the result conteined in the NIST MS Search file.

```{r }

# Función para parsear resultados NIST
parse_nist_msdial <- function(lines) {
  data_list <- list()
  current_unknown <- NA
  current_rt <- NA
  for (line in lines) {
    # Detectar feature
    if (grepl("^Unknown:", line)) {
      current_unknown <- str_extract(line, "UnknownID=\\d+") %>%
        str_extract("\\d+") %>%
        as.integer()
      current_rt <- str_extract(line, "RT=\\d+\\.?\\d*") %>%
        str_extract("\\d+\\.?\\d*") %>%
        as.numeric()
    }
    # Detectar hits
    if (grepl("^Hit", line)) {
      # Extract values from ";"
      parts <- unlist(str_split(line, ";"))
      parts <- trimws(parts)
      # Cleaning the values
      hit_num <- str_extract(line, "Hit\\s+\\d+") %>% str_extract("\\d+")
      name <- str_extract(parts[1], "<<.*?>>") %>% str_remove_all("<<|>>")
      formula <- str_extract(parts[2], "<<.*?>>") %>% str_remove_all("<<|>>")
      MF <- str_extract(parts[3], "\\d+")
      RMF <- str_extract(parts[4], "\\d+")
      Prob <- str_extract(parts[5], "-?\\d+\\.?\\d*")
      CAS <- str_extract(parts[6], "[0-9\\-]+")
      MW <- str_extract(parts[7], "\\d+")
      Lib <- str_extract(parts[8], "<<.*?>>") %>% str_remove_all("<<|>>")
      ID <- str_extract(parts[9], "\\d+")
      RI <- str_extract(parts[10], "\\d+")
      # Creating DataFrame
      row <- data.frame(
        Alignment_ID = current_unknown,
        RT = current_rt,
        Hit = as.integer(hit_num),
        Name = name,
        Formula = formula,
        MatchFactor = as.numeric(MF),
        ReverseMatch = as.numeric(RMF),
        Probability = as.numeric(Prob),
        CAS = CAS,
        MW = as.numeric(MW),
        Library = Lib,
        DB_ID = as.integer(ID),
        Lit_RI = as.numeric(RI),
        stringsAsFactors = FALSE
      )
      data_list[[length(data_list) + 1]] <- row
    }
  }
  bind_rows(data_list)
}
# Apply for each segment
md3t49m_df  <- parse_nist_msdial(md3t49m)
md49t54m_df <- parse_nist_msdial(md49t54m)
md54t56m_df <- parse_nist_msdial(md54t56m)
# Merge all
md_ids <- bind_rows(
  md3t49m_df %>% mutate(Segment = "3-49min"),
  md49t54m_df %>% mutate(Segment = "49-54min"),
  md54t56m_df %>% mutate(Segment = "54-56min"))
# Adding a global ID
md_ids <- md_ids %>%
  mutate(Feature_ID = paste0(Segment, "_", Alignment_ID))
# Moving columns
md_ids <- md_ids[,c(15, 14, 1:13)]

```

Loading the feature table with experimental RI.

```{r }

# Loading the feature list table with experimental RI
file_mdri <- "../X_hylaeae_metabolomics/Result/MSDIAL_feat_list_with_RI.xlsx"
md_ri <- read_excel(file_mdri, sheet = 1)
# Extracting only the experimental IR
md_ri <- data.frame(Segment = md_ri$Segment, Alignment_ID = md_ri$Alignment_ID,
                    Exp_RI = md_ri$R_RI, RI_tag = md_ri$RI_tag)

```

Adding the experimental RI to the DataFrame with identified metabolites.

```{r }

# Ensure the value format
md_ids$Alignment_ID <- as.character(md_ids$Alignment_ID)
md_ri$Alignment_ID <- as.character(md_ri$Alignment_ID)
# Adding experimental RI to DataFrame with identified metabolites
md_eri <- md_ids %>%
  left_join(md_ri, by = c("Segment", "Alignment_ID"))

```

Ading a column with a diference between literature RI and experimental RI.

```{r }

# Adding RI error
md_error <- md_eri %>%
  mutate(RI_Error = ifelse(
    Lit_RI == 0 | is.na(Lit_RI) | is.na(Exp_RI), NA, round(abs(Lit_RI - Exp_RI))))
# Moving columns
md_error <- md_error[,c(1:15, 17, 16)]

```

Adding a column that fag features with low match match (< 700).

```{r }

md_flag <- md_error %>%
  group_by(Feature_ID) %>%
  mutate(Match_flag = ifelse(MatchFactor < 700,
                             "Low score (<700)", "")) %>%
  ungroup()

```

Exporting the feature list with fag if its have match o no (Match factor below of 700).

```{r }

md_xp1 <- md_flag %>%
  group_by(Feature_ID) %>%
  # Evaluation condition by feature
  mutate(all_low = all(Match_flag == "Low score (<700)", na.rm = TRUE) &
           !any(is.na(Match_flag) | Match_flag == "")) %>%
  # Reduce a one row per feature
  slice(1) %>%
  # Apply condition over Match_factor
  mutate(MatchFactor = ifelse(all_low, MatchFactor, NA),
         Match_flag = ifelse(all_low, "Low score (<700)", NA)) %>%
  ungroup() %>%
  select(-all_low)
# Orderening by the Alignment_ID
md_xp1 <- md_xp1 %>%
  mutate(Alignment_ID = as.numeric(Alignment_ID)) %>%  # Change to numeric
  arrange(Alignment_ID)
# Exporting the feature list if its have match (match factor > 700)
md_xp1 %>%
  select(Feature_ID, Alignment_ID, RT, Match_flag) %>%
  write_xlsx("../X_hylaeae_metabolomics/Result/MSDIAL_feat_List_fg.xlsx")

```

**NOTE.** This step was used to quality control, the result of this was verify by manual identification.

Deleiting hits with low match.

```{r }

md3t49m_clean <- md3t49m_flag %>%
  group_by(Alignment_ID) %>%
  group_modify(~{
    if (all(.x$Match_flag == "Low score (<700)")) {
      .x %>%
        slice(1) %>%
        mutate(across(-c(RT, Exp_IR, Match_flag), ~ NA))
    } else {
      .x %>%
        filter(Match_flag != "Low score (<700)")
    }
  }) %>%
  ungroup()
# Orderening by the Alignment_ID
md3t49m_clean <- md3t49m_clean %>%
  mutate(Alignment_ID = as.numeric(Alignment_ID)) %>%  # Change to numeric
  arrange(Alignment_ID)
# Deleiting ionization tag in in-house library "[M]+.  70eV"
md3t49m_clean <- md3t49m_clean %>%
  mutate(
    Name = str_replace(Name, "\\s*\\[M\\]\\+\\.\\s*70\\s*eV", ""))

```

Now, we focus on feature with match with difference between experimental RI and literature RI between 0 and 10. In the case that the feature meet this condition, only match with high match score will be reported. If you have a in house library made using autentic reference standard you can prioritize this library in order to report level 1 identification level.

Now we focus in level 3 identification level. Here we separate in two filter, one features between 10 and 100 of RI desviation, and feature that have more of 100 RI desviation or its do not have RI desviation. In the fist case, we prioritize the low RI desviation, and in the second case we prioritaze the high match factor.

```{r }

md3t49m_il <- md3t49m_clean %>%
  group_by(Alignment_ID) %>%
  mutate(
    # Clasificación por rangos
    in_0_10   = !is.na(RI_Error) & RI_Error <= 10,
    in_10_100 = !is.na(RI_Error) & RI_Error > 10 & RI_Error <= 100,
    has_0_10   = any(in_0_10),
    has_10_100 = any(in_10_100),
    is_ikiam = Library == "ikiam_npl_gcms_lib") %>%
  # Asignar prioridad por fila
  mutate(priority = case_when(
      has_0_10   & in_0_10   ~ 1,
      !has_0_10 & has_10_100 & in_10_100 ~ 2,
      !has_0_10 & !has_10_100 ~ 3,
      TRUE ~ NA_real_)) %>%
  # Mantener solo filas relevantes por feature
  filter(!is.na(priority)) %>%
  # Orden según prioridad
  arrange(priority,
    # PRIORIDAD 1 → Match alto primero o in-house library si aplica
    desc(ifelse(priority == 1, is_ikiam, FALSE)),
    desc(ifelse(priority == 1, MatchFactor, NA)),
    ifelse(priority == 1, RI_Error, NA),
    # PRIORIDAD 2 → RI bajo primero
    ifelse(priority == 2, RI_Error, NA),
    desc(ifelse(priority == 2, MatchFactor, NA)),
    # PRIORIDAD 3 → Match alto
    desc(ifelse(priority == 3, MatchFactor, NA))) %>%
  slice(1) %>%
  ungroup() %>%
  select(-in_0_10, -in_10_100, -has_0_10, -has_10_100, -priority, -is_ikiam)
# Orderening by the Alignment_ID
md3t49m_il <- md3t49m_il %>%
  mutate(Alignment_ID = as.numeric(Alignment_ID)) %>%  # Change to numeric
  arrange(Alignment_ID)

```

En some case same name of metabolite has been asigned for multiple feature. In the case with same metabolite name was assigned for multiple feature, the best match factor was priorized for features between 0 to 10 RI error (the in hose library was priorized a them the filter was apply). For features with same metabolite name and RI error between 10 and 100, the metabolie name was assigned for feature with the low RI error, in the case with the same RI error the best match factor was priorized. For features with the same metabolite name and RI error > 100 or that have NA, the metabolite name was assigned to feature with major match factor.

```{r }

# Separar identificados
md3t49m_ids <- md3t49m_il %>%
  filter(!is.na(Name))
# Separar no identificados
md3t49m_noids <- md3t49m_il %>%
  filter(is.na(Name))
# Aplicar lógica solo a identificados
md3t49m_rep <- md3t49m_ids %>%   # Tabla con 1 hit por feature
  group_by(Name) %>%
  mutate(in_0_10   = !is.na(RI_Error) & RI_Error <= 10,
         in_10_100 = !is.na(RI_Error) & RI_Error > 10 & RI_Error <= 100,
         has_0_10   = any(in_0_10),
         has_10_100 = any(in_10_100),
         is_ikiam = Library == "ikiam_npl_gcms_lib") %>%
  # Definir grupo activo
  mutate(active_group = case_when(
    has_0_10 ~ "0_10",
    !has_0_10 & has_10_100 ~ "10_100",
    TRUE ~ "gt100")) %>%
  # Definir quién participa en el ranking
  mutate(candidate = case_when(active_group == "0_10"   ~ in_0_10,
                               active_group == "10_100" ~ in_10_100,
                               TRUE ~ TRUE)) %>%
  # Orden SOLO para candidatos
  arrange(Name,
    desc(candidate),  # candidatos primero
    # PRIORIDAD 1
    desc(ifelse(candidate & active_group == "0_10", is_ikiam, FALSE)),
    desc(ifelse(candidate & active_group == "0_10", MatchFactor, NA)),
    ifelse(candidate & active_group == "0_10", RI_Error, Inf),
    # PRIORIDAD 2
    ifelse(candidate & active_group == "10_100", RI_Error, Inf),
    desc(ifelse(candidate & active_group == "10_100", MatchFactor, NA)),
    # PRIORIDAD 3
    desc(ifelse(candidate & active_group == "gt100", MatchFactor, NA))) %>%
  group_by(Name) %>%
  mutate(rank_name = ifelse(candidate, row_number(), NA_integer_)) %>%
  ungroup() %>%
  # Etiquetado
  mutate(Match_flag = case_when(
    rank_name == 1 ~ Match_flag,
    active_group == "0_10" ~ "Repeated (RI error <10)",
    active_group == "10_100" ~ "Repeated (RI error 10-100)",
    TRUE ~ "Repeated (RI error >100/NA)")) %>%
  select(-in_0_10, -in_10_100, -has_0_10, -has_10_100,
         -is_ikiam, -active_group, -candidate, -rank_name)
# Volver a unir todo
md3t49m_rep <- bind_rows(md3t49m_rep, md3t49m_noids)
# Orderening by the Alignment_ID
md3t49m_rep <- md3t49m_rep %>%
  mutate(Alignment_ID = as.numeric(Alignment_ID)) %>%  # Change to numeric
  arrange(Alignment_ID)

```

Adding a identification level, metabolites identified with in house library (match factor > 700) and RI error < 10, will be tag as level 1 identification confidence level, metabolites identified with NIST libraries (match factor > 700) and RI error < 10, wil be tag as level 2 identification confidence level. Metabolites identified wit NIST libraries (match factor > 700) and RI error > 10 was assigned as level 3 identification confidense level.

```{r }

md3t49m_il123 <- md3t49m_rep %>%
  mutate(IL = case_when(
      # No etiquetar
      is.na(Name) ~ NA_character_,
      Match_flag == "Low score (<700)" ~ NA_character_,
      # Level 1
      !is.na(RI_Error) & RI_Error <= 10 & 
        Library == "ikiam_npl_gcms_lib" ~ "Level 1",
      # Level 2
      !is.na(RI_Error) & RI_Error <= 10 ~ "Level 2",
      # Level 3
      TRUE ~ "Level 3"))

```

Now we will use the NIST data base (mainlib and replib) to found a InChIKey, and short name to add to the identified table.

```{r }

# Función para parsear un archivo NIST .txt
parse_nist_txt <- function(file_path) {
  lines <- readLines(file_path, warn = FALSE)
  # Separar por compuestos
  blocks <- split(lines, cumsum(grepl("^Name:", lines)))
  parse_block <- function(block) {
    block_txt <- paste(block, collapse = "\n")
    # Campos principales
    name <- str_match(block_txt, "Name:\\s*(.*)")[,2]
    inchikey <- str_match(block_txt, "InChIKey:\\s*(.*)")[,2]
    # Synon (hasta 3)
    synon_all <- str_match_all(block_txt, "Synon:\\s*(.*)")[[1]][,2]
    synon1 <- if(length(synon_all) >= 1) synon_all[1] else NA
    synon2 <- if(length(synon_all) >= 2) synon_all[2] else NA
    synon3 <- if(length(synon_all) >= 3) synon_all[3] else NA
    # CAS y Related CAS
    cas <- str_match(block_txt, "CAS#:\\s*([^;\\n]+)")[,2]
    related_cas <- str_match(block_txt, "Related_CAS#:\\s*([^;\\n]+)")[,2]
    # NIST ID
    nist_id <- str_match(block_txt, "NIST#:\\s*(\\d+)")[,2]
    # DB#
    db_id <- str_match(block_txt, "DB#:\\s*(\\d+)")[,2]
    # Retention Index
    ri_line <- str_match(block_txt, "Retention_index:\\s*(.*)")[,2]
    ri_semi <- NA
    ri_stdnp <- NA
    ri_polar <- NA
    if(!is.na(ri_line)) {
      parts <- str_split(ri_line, "\\s+")[[1]]
      for(p in parts) {
        # SemiStdNP
        if(str_detect(p, "SemiStdNP=")) {
          val <- str_match(p, "SemiStdNP=(\\d+)")[,2]
          if(is.na(val)) val <- str_match(p, "SemiStdNP=(\\d+)/")[,2]
          ri_semi <- as.numeric(val)}
        # StdNP
        if(str_detect(p, "StdNP=")) {
          val <- str_match(p, "StdNP=(\\d+)")[,2]
          if(is.na(val)) val <- str_match(p, "StdNP=(\\d+)/")[,2]
          ri_stdnp <- as.numeric(val)}
        # StdPolar
        if(str_detect(p, "StdPolar=")) {
          val <- str_match(p, "StdPolar=(\\d+)")[,2]
          if(is.na(val)) val <- str_match(p, "StdPolar=(\\d+)/")[,2]
          ri_polar <- as.numeric(val)
        }
      }
    }
    tibble(
      Name = name,
      Synon1 = synon1,
      Synon2 = synon2,
      Synon3 = synon3,
      InChIKey = inchikey,
      CAS = cas,
      Related_CAS = related_cas,
      NIST_ID = nist_id,
      DB_ID = db_id,
      RI_SemiStdNP = ri_semi,
      RI_StdNP = ri_stdnp,
      RI_StdPolar = ri_polar
    )
  }
  map_dfr(blocks, parse_block)
}
# Cargar bibliotecas
mainlib <- parse_nist_txt("F:/NIST_20_Library/Data/mainlib.txt")
replib  <- parse_nist_txt("F:/NIST_20_Library/Data/replib.txt")
ikiamlib  <- parse_nist_txt("F:/NIST_20_Library/Data/ikiamlib.txt")
# Unir
nist_db <- bind_rows(
  mainlib %>% mutate(Library = "mainlib"),
  replib  %>% mutate(Library = "replib"),
  ikiamlib %>% mutate(Library = "ikiam_npl_gcms_lib"))
# Ensuring a unic compound in the data base
nist_clean <- nist_db %>%
  distinct(DB_ID, Library, .keep_all = TRUE)
# Saving the processed database
saveRDS(nist_clean, "../X_hylaeae_metabolomics/Result/nist_db.rds")

```

Using a NIST 20 library the InChIKey was added to the identified table, the data base ID and library of identified metabolites was used to found the correct InChIKey.

```{r }

md3t49m_ik <- md3t49m_il123 %>%
  mutate(DB_ID = as.character(DB_ID)) %>%
  left_join(nist_clean %>%
              mutate(DB_ID = as.character(DB_ID)) %>%
              select(DB_ID, Library, InChIKey),
            by = c("DB_ID", "Library"))

```


































