---
title: "Metabolite identified using NIST MS Search (NIST 20: *mainlib* and *replib* libraries) with data deconvoluted using MS-DIAL - SH-Rtx-Wax Polar colum"
author: "Jefferson Pastuna"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
usethis::git_vaccinate()

```

## Introduction

This notebook aim show how the hits result from NIST MS Search using NIST 20 library was filtered in order tu report metabolites with the best match and small RI shift.

### MS-DIAL

Loading the necesary libraries.

```{r }

library(stringr)
library(dplyr)
library(tidyr)
library(readxl)
library(writexl)
library(purrr)
library(classyfireR)

```

Reading the files.

```{r }

# Loading identification files exported from NIST MS Search
file_md3t38m_wx <- "../X_hylaeae_metabolomics/Data/NIST_MS_Search/MSDIAL_GCMS_polar_3-38min.TXT"
file_md38t50m_wx <- "../X_hylaeae_metabolomics/Data/NIST_MS_Search/MSDIAL_GCMS_polar_38-50min.TXT"
md3t38m_wx <- readLines(file_md3t38m_wx, encoding = "latin1")
md38t50m_wx <- readLines(file_md38t50m_wx, encoding = "latin1")

```

Formating the result conteined in the NIST MS Search file.

```{r }

# Fumction to extract NIST result
parse_nist_msdial <- function(lines) {
  data_list <- list()
  current_unknown <- NA
  current_rt <- NA
  for (line in lines) {
    # Feature detection
    if (str_detect(line, "^Unknown:")) {
      current_unknown <- str_extract(line, "UnknownID=\\d+") %>%
        str_extract("\\d+") %>%
        as.integer()
      current_rt <- str_extract(line, "RT=\\d+\\.?\\d*") %>%
        str_extract("\\d+\\.?\\d*") %>%
        as.numeric()
    }
    # Hits detection
    if (str_detect(line, "^Hit")) {
      # Character detection << >> and extract by block
      blocks <- str_extract_all(line, "<<.*?>>")[[1]]
      blocks <- str_remove_all(blocks, "<<|>>")
      name    <- ifelse(length(blocks) >= 1, blocks[1], NA)
      formula <- ifelse(length(blocks) >= 2, blocks[2], NA)
      library <- ifelse(length(blocks) >= 3, blocks[3], NA)
      # Value extraction
      hit_num <- str_extract(line, "Hit\\s+\\d+") %>% str_extract("\\d+")
      MF  <- str_extract(line, "MF:\\s*\\d+") %>% str_extract("\\d+")
      RMF <- str_extract(line, "RMF:\\s*\\d+") %>% str_extract("\\d+")
      Prob <- str_extract(line, "Prob:\\s*-?\\d+\\.?\\d*") %>% str_extract("-?\\d+\\.?\\d*")
      CAS <- str_extract(line, "CAS:\\s*[0-9\\-]+") %>% str_extract("[0-9\\-]+")
      MW  <- str_extract(line, "Mw:\\s*\\d+") %>% str_extract("\\d+")
      ID  <- str_extract(line, "Id:\\s*\\d+") %>% str_extract("\\d+")
      RI  <- str_extract(line, "RI:\\s*\\d+") %>% str_extract("\\d+")
      # Adding values by row
      row <- data.frame(
        Alignment_ID = current_unknown,
        RT = current_rt,
        Hit = as.integer(hit_num),
        Name = name,
        Formula = formula,
        MatchFactor = as.numeric(MF),
        ReverseMatch = as.numeric(RMF),
        Probability = as.numeric(Prob),
        CAS = CAS,
        MW = as.numeric(MW),
        Library = library,
        DB_ID = as.integer(ID),
        Lit_RI = as.numeric(RI),
        stringsAsFactors = FALSE
      )
      data_list[[length(data_list) + 1]] <- row
    }
  }
  bind_rows(data_list)
}
# Apply for each segment
md3t38m_wx_df  <- parse_nist_msdial(md3t38m_wx)
md38t50m_wx_df <- parse_nist_msdial(md38t50m_wx)
# Merge all
md_wx_ids <- bind_rows(
  md3t38m_wx_df %>% mutate(Segment = "3-38min"),
  md38t50m_wx_df %>% mutate(Segment = "38-50min"))
# Moving columns
md_wx_ids <- md_wx_ids[,c(14, 1:13)]

```

Loading the feature table with experimental RI.

```{r }

# Loading the feature list table with experimental RI
file_mdriwx <- "../X_hylaeae_metabolomics/Result/MSDIAL_GCMS_polar_feat_list_with_RI.xlsx"
md_wx_ri <- read_excel(file_mdriwx, sheet = 1)
# Extracting only the experimental IR
md_wx_ri <- data.frame(Feature_ID = md_wx_ri$Feature_ID, Segment = md_wx_ri$Segment,
                       Alignment_ID = md_wx_ri$Alignment_ID, Exp_RI = md_wx_ri$R_RI,
                       RI_tag = md_wx_ri$RI_tag)

```

Adding the experimental RI to the DataFrame with identified metabolites.

```{r }

# Ensure the value format
md_wx_ids$Alignment_ID <- as.character(md_wx_ids$Alignment_ID)
md_wx_ri$Alignment_ID <- as.character(md_wx_ri$Alignment_ID)
# Adding experimental RI to DataFrame with identified metabolites
md_wx_eri <- md_wx_ids %>%
  left_join(md_wx_ri, by = c("Segment", "Alignment_ID"))

```

Adding the NIST metadata (InChIKey, Synonyms, and Polar RI) of identified metabolites.

```{r }

# Load the NIST database (metadata information of library)
nist_db <- readRDS("F:/NIST_20_Library/Result/nist_db.rds")
# Adding aditional information to the identified metabolites
md_wx_lri <- md_wx_eri %>%
  mutate(DB_ID = as.character(DB_ID)) %>%
  left_join(nist_db %>%
              mutate(DB_ID = as.character(DB_ID)) %>%
              select(DB_ID, Library, RI_StdPolar, InChIKey, Synon1, Synon2, Synon3),
            by = c("DB_ID", "Library"))

```

Ading a column with a diference between literature RI and experimental RI.

```{r }

# Adding RI error
md_wx_error <- md_wx_lri %>%
  mutate(RI_Error = ifelse(
    RI_StdPolar == 0 | is.na(RI_StdPolar) | is.na(Exp_RI), NA, round(abs(RI_StdPolar - Exp_RI))))
# Moving columns
md_wx_error <- md_wx_error[,c(15, 1:14, 16, 18, 23, 19:22, 17)]

```

Adding a column that fag features with low match match (< 700).

```{r }

md_wx_flag <- md_wx_error %>%
  group_by(Feature_ID) %>%
  mutate(Match_flag = ifelse(MatchFactor < 700,
                             "Low score (<700)", "")) %>%
  ungroup()

```

Deleiting hits with low match.

```{r }

md_wx_clean <- md_wx_flag %>%
  group_by(Feature_ID) %>%
  group_modify(~{
    if (all(.x$Match_flag == "Low score (<700)")) {
      .x %>%
        slice(1) %>%
        mutate(across(-c(Segment, Alignment_ID, RT, Exp_RI, Match_flag), ~ NA))
    } else {
      .x %>%
        filter(Match_flag != "Low score (<700)")
    }
  }) %>%
  ungroup()
# Orderening by the Alignment_ID
md_wx_clean <- md_wx_clean %>%
  mutate(Feature_ID = as.numeric(Feature_ID)) %>%  # Change to numeric
  arrange(Feature_ID)
# Deleting ionization tag in the name of in-house library "[M]+.  70eV"
md_wx_clean <- md_wx_clean %>%
  mutate(
    Name = str_replace(Name, "\\s*\\[M\\]\\+\\.\\s*70\\s*eV", ""))

```

Now, we focus on feature with match with difference between experimental RI and literature RI between 0 and 10. In the case that the feature meet this condition, only match with high match score will be reported. If you have a in house library made using autentic reference standard you can prioritize this library in order to report level 1 identification level.

Now we focus in level 3 identification level. Here we separate in two filter, one features between 10 and 100 of RI desviation, and feature that have more of 100 RI desviation or its do not have RI desviation. In the fist case, we prioritize the low RI desviation, and in the second case we prioritaze the high match factor.

```{r }

md_wx_il <- md_wx_clean %>%
  group_by(Feature_ID) %>%
  mutate(
    # Classification by rage
    in_0_10   = !is.na(RI_Error) & RI_Error <= 10,
    in_10_100 = !is.na(RI_Error) & RI_Error > 10 & RI_Error <= 100,
    has_0_10   = any(in_0_10),
    has_10_100 = any(in_10_100),
    is_ikiam = Library == "ikiam_npl_gcms_lib") %>%
  # Adding priority by row
  mutate(priority = case_when(
      has_0_10   & in_0_10   ~ 1,
      !has_0_10 & has_10_100 & in_10_100 ~ 2,
      !has_0_10 & !has_10_100 ~ 3,
      TRUE ~ NA_real_)) %>%
  # Keep only key row by each row
  filter(!is.na(priority)) %>%
  # ordering according the priority
  arrange(priority,
    # Priority 1: High match first or in house library of apply
    desc(ifelse(priority == 1, is_ikiam, FALSE)),
    desc(ifelse(priority == 1, MatchFactor, NA)),
    ifelse(priority == 1, RI_Error, NA),
    # Priority 2: Low RI first
    ifelse(priority == 2, RI_Error, NA),
    desc(ifelse(priority == 2, MatchFactor, NA)),
    # Priority 3: High match first
    desc(ifelse(priority == 3, MatchFactor, NA))) %>%
  slice(1) %>%
  ungroup() %>%
  select(-in_0_10, -in_10_100, -has_0_10, -has_10_100, -priority, -is_ikiam)
# Orderening by the Feature_ID
md_wx_il <- md_wx_il %>%
  mutate(Feature_ID = as.numeric(Feature_ID)) %>%  # Change to numeric
  arrange(Feature_ID)

```

En some case same name of metabolite has been asigned for multiple feature. In the case with same metabolite name was assigned for multiple feature, the best match factor was priorized for features between 0 to 10 RI error (the in hose library was priorized a them the filter was apply). For features with same metabolite name and RI error between 10 and 100, the metabolie name was assigned for feature with the low RI error, in the case with the same RI error the best match factor was priorized. For features with the same metabolite name and RI error > 100 or that have NA, the metabolite name was assigned to feature with major match factor. Feature that not meet the condition descrive above was tag as "Repeated (RI error <10)", "Repeated (RI error 10-100)", and "Repeated (RI error >100/NA)", accordingly.

```{r }

# Extracting features with IDs
md_wx_ids <- md_wx_il %>%
  filter(!is.na(Name))
# Extracting features without IDs
md_wx_noids <- md_wx_il %>%
  filter(is.na(Name))
# Applying firter only for features with IDs
md_wx_rep <- md_wx_ids %>%   # Table with 1 hit per feature
  group_by(Name) %>%
  mutate(in_0_10   = !is.na(RI_Error) & RI_Error <= 10,
         in_10_100 = !is.na(RI_Error) & RI_Error > 10 & RI_Error <= 100,
         has_0_10   = any(in_0_10),
         has_10_100 = any(in_10_100),
         is_ikiam = Library == "ikiam_npl_gcms_lib") %>%
  # Select active group
  mutate(active_group = case_when(
    has_0_10 ~ "0_10",
    !has_0_10 & has_10_100 ~ "10_100",
    TRUE ~ "gt100")) %>%
  # Define who participates in the ranking
  mutate(candidate = case_when(active_group == "0_10"   ~ in_0_10,
                               active_group == "10_100" ~ in_10_100,
                               TRUE ~ TRUE)) %>%
  # Order ONLY for candidates
  arrange(Name,
    desc(candidate),  # candidates first
    # PRIORITY 1
    desc(ifelse(candidate & active_group == "0_10", is_ikiam, FALSE)),
    desc(ifelse(candidate & active_group == "0_10", MatchFactor, NA)),
    ifelse(candidate & active_group == "0_10", RI_Error, Inf),
    # PRIORITY 2
    ifelse(candidate & active_group == "10_100", RI_Error, Inf),
    desc(ifelse(candidate & active_group == "10_100", MatchFactor, NA)),
    # PRIORITY 3
    desc(ifelse(candidate & active_group == "gt100", MatchFactor, NA))) %>%
  group_by(Name) %>%
  mutate(rank_name = ifelse(candidate, row_number(), NA_integer_)) %>%
  ungroup() %>%
  # Labeled
  mutate(Match_flag = case_when(
    rank_name == 1 ~ Match_flag,
    active_group == "0_10" ~ "Repeated (RI error <10)",
    active_group == "10_100" ~ "Repeated (RI error 10-100)",
    TRUE ~ "Repeated (RI error >100/NA)")) %>%
  select(-in_0_10, -in_10_100, -has_0_10, -has_10_100,
         -is_ikiam, -active_group, -candidate, -rank_name)
# Merge all
md_wx_rep <- bind_rows(md_wx_rep, md_wx_noids)
# Orderening by the Feature_ID
md_wx_rep <- md_wx_rep %>%
  mutate(Feature_ID = as.numeric(Feature_ID)) %>%  # Change to numeric
  arrange(Feature_ID)

```

Adding a identification level, metabolites identified with in house library (match factor > 700) and RI error < 10, will be tag as level 1 identification confidence level, metabolites identified with NIST libraries (match factor > 700) and RI error < 10, wil be tag as level 2 identification confidence level. Metabolites identified wit NIST libraries (match factor > 700) and RI error > 10 was assigned as level 3 identification confidense level.

```{r }

md_wx_il123 <- md_wx_rep %>%
  mutate(IL = case_when(
      # No etiquetar
      is.na(Name) ~ NA_character_,
      Match_flag == "Low score (<700)" ~ NA_character_,
      # Level 1
      !is.na(RI_Error) & RI_Error <= 10 & 
        Library == "ikiam_npl_gcms_lib" ~ "Level 1",
      # Level 2
      !is.na(RI_Error) & RI_Error <= 10 ~ "Level 2",
      # Level 3
      TRUE ~ "Level 3"))

```

Adding espectrum data from the NIST libry to the identified metabolites.

```{r }
# Load the NIST database (spectrum information of library)
nist_mz <- readRDS("F:/NIST_20_Library/Result/nist_mz.rds")
# Adding espectrum data from the NIST libry to the identified metabolites
md_wx_libmz <- md_wx_il123 %>%
  mutate(DB_ID = as.character(DB_ID)) %>%
  left_join(nist_mz %>%
              mutate(DB_ID = as.character(DB_ID)) %>%
              select(DB_ID, Library, nist_mz),
            by = c("DB_ID", "Library"))

```

Now we going to add the experimental EI spectrum to the identified feature, in the following steps we could compare the spectrum between NIST spectrum and experimental spectrum.

```{r }

# Funtion to separate ions and intensity from experimental EI spectrum
parse_exp_ei <- function(spec_string) {
  if(is.na(spec_string)) return(list(mz = NA, int = NA))
  pairs <- str_split(spec_string, " ")[[1]]
  mz <- as.numeric(str_extract(pairs, "^[0-9\\.]+"))
  intensity <- as.numeric(str_extract(pairs, "(?<=:)\\d+"))
  list(Exp_mz = mz, Exp_int = intensity)
}
# Loading experimental spectrum
file_mdeiwx <- "../X_hylaeae_metabolomics/Result/MSDIAL_GCMS_polar_feat_list_with_RI_EI.xlsx"
md_eiwx <- read_excel(file_mdeiwx, sheet = 1)
# Extracting only the experimental EI spectrum
md_eiwx <- data.frame(Feature_ID = md_eiwx$Feature_ID, Exp_EI = md_eiwx$Exp_EI)
# Addin experimental spectum to edentified metabolites
md_wx_expmz <- md_wx_libmz %>%
  left_join(md_eiwx %>%
              select(Feature_ID, Exp_EI), by = "Feature_ID")
# Change the spectrum to list
md_wx_expmz_ls <- md_wx_expmz %>%
  rowwise() %>%
  mutate(spec = list(parse_exp_ei(Exp_EI)),
         exp_mz = list(spec$Exp_mz),
         exp_int = list(spec$Exp_int)) %>%
  ungroup() %>%
  select(-spec)
# Adding intesity filter (nois) and pass to nominal mass
md_wx_expmz_ns <- md_wx_expmz_ls %>%
  rowwise() %>%
  mutate(int_exp_norm = list(exp_int / max(exp_int, na.rm = TRUE) * 100),
         keep_idx = list(which(int_exp_norm >= 2)),
         exp_mz_ns = list(round(exp_mz[keep_idx])),
         exp_int_ns = list(int_exp_norm[keep_idx])) %>%
  ungroup()

```

The folowing code count the number of ions are equal between library and experiment spectrum.

```{r }

# Compare to only identified features
md_wx_expmz_id <- md_wx_expmz_ns %>%
  filter(!is.na(Name)) %>%
  rowwise() %>%
  mutate(matched_ions = length(intersect(exp_mz_ns, nist_mz))) %>%
  ungroup()
# Extracting no identified features
md_wx_expmz_noid <- md_wx_expmz_ns %>%
  filter(is.na(Name)) %>%
  mutate(matched_ions = NA_integer_)
# Join identified and no identified features with number of ion match
md_wx_ions <- bind_rows(md_wx_expmz_id, md_wx_expmz_noid)
# Orderening by the Feature_ID
md_wx_ions <- md_wx_ions %>%
  mutate(Feature_ID = as.numeric(Feature_ID)) %>%  # Change to numeric
  arrange(Feature_ID)

```

Adding a flag if the number of ions match is below of 5.

```{r }

md_wx_ions_fg <- md_wx_ions %>%
  mutate(Match_flag = case_when(
    # Unidentified (do not edit)
    is.na(Name) ~ Match_flag,
    # Low ion match
    !is.na(matched_ions) & matched_ions < 5 ~ 
      ifelse(is.na(Match_flag), "Low ion match (<5)",
             paste(Match_flag, "Low ion match (<5)", sep = "; ")),
    # Keep the rest
    TRUE ~ Match_flag))

```

Flag features identified as siloxane compounds by compund formula.

```{r }

md_wx_ions_si <- md_wx_ions_fg %>%
  mutate(Match_flag = case_when(
    !is.na(Formula) & str_detect(Formula, "Si\\d*") ~
      ifelse(is.na(Match_flag),
             "Siloxane",
             paste(Match_flag, "Siloxane", sep = "; ")),
    TRUE ~ Match_flag))

```

Extracting InChIKey for classification of type of molecule using a Classyfire tool (https://github.com/aberHRML/classyfireR).

```{r }

# Extractiong the InChIKey from the feature file
md_ions_ik <- md_ions_si %>%
  group_by(InChIKey) %>%
  filter(any(is.na(Match_flag) | Match_flag == "")) %>%
  ungroup() %>%
  filter(!is.na(InChIKey)) %>%
  distinct(InChIKey)
# Metabolite classification
md_wx_met_class <- purrr::map_dfr(md_wx_ions_ik$InChIKey, function(ik) {
  res <- tryCatch(get_classification(ik), error = function(e) NULL)
  if (is.null(res) || is.null(res@classification)) {
    return(tibble(
      InChIKey = ik,
      Superclass = NA,
      Class = NA,
      Subclass = NA
    ))
  }
  class_vec <- res@classification$Classification
  # Validar vector
  if (!is.character(class_vec) || length(class_vec) < 2) {
    return(tibble(
      InChIKey = ik,
      Superclass = NA,
      Class = NA,
      Subclass = NA
    ))
  }
  tibble(
    InChIKey = ik,
    Superclass = ifelse(length(class_vec) >= 2, class_vec[2], NA),
    Class      = ifelse(length(class_vec) >= 3, class_vec[3], NA),
    Subclass   = ifelse(length(class_vec) >= 4, class_vec[4], NA)
  )
})

```

Adding a classification of moleculecules to the feature table of identified features.

```{r }

# Join with original table the molecules classification
md_wx_ions_cf <- md_wx_ions_si %>%
  left_join(md_wx_met_class, by = "InChIKey")
# Exporting the feature list with metabolites identified at IL 1, 2, and 3
md_wx_ions_cf %>%
  select(Feature_ID, Segment, Alignment_ID, IL, RT, Name, Synon1, Synon2, Synon3,
         Formula, MatchFactor, matched_ions, RI_StdPolar, Exp_RI, RI_Error, Library,
         DB_ID, InChIKey, Superclass, Class, Subclass,Match_flag, RI_tag) %>%
  write_xlsx("../X_hylaeae_metabolomics/Result/MSDIAL_GCMS_polar_IDs.xlsx")

```


C:\Users\Waters>F:\NIST20\MSSEARCH\NISTMS$.EXE /instrument /PAR=2



