---
title: "Metabolite identified using NIST MS Search (NIST 20 *mainlib* and *replib* libraries) with data deconvoluted using MZmine - Rtx-5 Semi-standard non-polar colum"
author: "Jefferson Pastuna"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
usethis::git_vaccinate()

```

## Introduction

This notebook aim show how the hits result from NIST MS Search using NIST 20 library was filtered in order tu report metabolites with the best match and small RI shift.

### MS-DIAL

Loading the necesary libraries.

```{r }

library(stringr)
library(dplyr)
library(tidyr)
library(readxl)
library(writexl)
library(purrr)
library(classyfireR)

```

Reading the files.

```{r }

file_mn3t49m <- "../X_hylaeae_metabolomics/Data/NIST_MS_Search/MZmine_GCMS_apolar_3-49min.TXT"
file_mn49t54m <- "../X_hylaeae_metabolomics/Data/NIST_MS_Search/MZmine_GCMS_apolar_49-54min.TXT"
file_mn54t56m <- "../X_hylaeae_metabolomics/Data/NIST_MS_Search/MZmine_GCMS_apolar_54-56min.TXT"
mn3t49m <- readLines(file_mn3t49m, encoding = "latin1")
mn49t54m <- readLines(file_mn49t54m, encoding = "latin1")
mn54t56m <- readLines(file_mn54t56m, encoding = "latin1")

```

Formating the result conteined in the NIST MS Search file.

```{r }

# Fumction to extract NIST result
parse_nist_mzmine <- function(lines) {
  data_list <- list()
  current_unknown <- NA
  for (line in lines) {
    # Feature detection
    if (str_detect(line, "^Unknown:")) {
      current_unknown <- str_extract(line, "(?i)offset\\s*=\\s*(0x[0-9A-Fa-f]+|\\d+)") %>%
        str_extract("(0x[0-9A-Fa-f]+|\\d+)")
    }
    # Hits detection
    if (str_detect(line, "^Hit")) {
      # Character detection << >> and extract by block
      blocks <- str_extract_all(line, "<<.*?>>")[[1]]
      blocks <- str_remove_all(blocks, "<<|>>")
      name    <- ifelse(length(blocks) >= 1, blocks[1], NA)
      formula <- ifelse(length(blocks) >= 2, blocks[2], NA)
      library <- ifelse(length(blocks) >= 3, blocks[3], NA)
      # Value extraction
      hit_num <- str_extract(line, "Hit\\s+\\d+") %>% str_extract("\\d+")
      MF  <- str_extract(line, "MF:\\s*\\d+") %>% str_extract("\\d+")
      RMF <- str_extract(line, "RMF:\\s*\\d+") %>% str_extract("\\d+")
      Prob <- str_extract(line, "Prob:\\s*-?\\d+\\.?\\d*") %>% str_extract("-?\\d+\\.?\\d*")
      CAS <- str_extract(line, "CAS:\\s*[0-9\\-]+") %>% str_extract("[0-9\\-]+")
      MW  <- str_extract(line, "Mw:\\s*\\d+") %>% str_extract("\\d+")
      ID  <- str_extract(line, "Id:\\s*\\d+") %>% str_extract("\\d+")
      RI  <- str_extract(line, "RI:\\s*\\d+") %>% str_extract("\\d+")
      # Adding values by row
      row <- data.frame(
        NIST_Code = current_unknown,
        Hit = as.integer(hit_num),
        Name = name,
        Formula = formula,
        MatchFactor = as.numeric(MF),
        ReverseMatch = as.numeric(RMF),
        Probability = as.numeric(Prob),
        CAS = CAS,
        MW = as.numeric(MW),
        Library = library,
        DB_ID = as.integer(ID),
        Lit_RI = as.numeric(RI),
        stringsAsFactors = FALSE
      )
      data_list[[length(data_list) + 1]] <- row
    }
  }
  bind_rows(data_list)
}
# Apply for each segment
mn3t49m_df  <- parse_nist_mzmine(mn3t49m)
mn49t54m_df <- parse_nist_mzmine(mn49t54m)
mn54t56m_df <- parse_nist_mzmine(mn54t56m)
# Merge all
mn_ids <- bind_rows(
  mn3t49m_df %>% mutate(Segment = "3-49min"),
  mn49t54m_df %>% mutate(Segment = "49-54min"),
  mn54t56m_df %>% mutate(Segment = "54-56min"))%>%
  # Adding a global ID
  mutate(feature_key = paste(Segment, NIST_Code, sep = "_"),
         Feature_ID = dense_rank(feature_key)) %>%
  select(-feature_key)
# Moving columns
mn_ids <- mn_ids[,c(14, 13, 1:12)]

```

Loading the feature table with experimental RI.

```{r }

# Loading the feature list table with experimental RI
file_mnri <- "../X_hylaeae_metabolomics/Result/MZmine_GCMS_apolar_feat_list_with_RI.xlsx"
mn_ri <- read_excel(file_mnri, sheet = 1)
# Extracting only the experimental IR
mn_ri <- data.frame(Feature_ID = mn_ri$Feature_ID, Segment = mn_ri$Segment,
                    Alignment_ID = mn_ri$Alignment_ID, RT = mn_ri$RT,
                    Exp_RI = mn_ri$R_RI, RI_tag = mn_ri$RI_tag)

```

Adding the experimental RI to the DataFrame with identified metabolites.

```{r }

# Ensure the value format
mn_ids$Feature_ID <- as.character(mn_ids$Feature_ID)
mn_ri$Feature_ID <- as.character(mn_ri$Feature_ID)
# Adding experimental RI to DataFrame with identified metabolites
mn_eri <- mn_ids %>%
  left_join(mn_ri, by = c("Segment", "Feature_ID"))

```

Adding the NIST metadata (InChIKey, Synonyms, and Polar RI) of identified metabolites.

```{r }

# Load the NIST database (metadata information of library)
nist_db <- readRDS("F:/NIST_20_Library/Result/nist_db.rds")
# Adding aditional information to the identified metabolites
mn_ik <- mn_eri %>%
  mutate(DB_ID = as.character(DB_ID)) %>%
  left_join(nist_db %>%
              mutate(DB_ID = as.character(DB_ID)) %>%
              select(DB_ID, Library, InChIKey, Synon1, Synon2, Synon3),
            by = c("DB_ID", "Library"))

```

Ading a column with a diference between literature RI and experimental RI.

```{r }

# Adding RI error
mn_error <- mn_ik %>%
  mutate(RI_Error = ifelse(
    Lit_RI == 0 | is.na(Lit_RI) | is.na(Exp_RI), NA, round(abs(Lit_RI - Exp_RI))))
# Moving columns
mn_error <- mn_error[,c(1, 15, 2:13, 16, 14, 17, 23, 18, 19:22)]

```

Adding a column that fag features with low match match (< 700).

```{r }

mn_flag <- mn_error %>%
  group_by(Feature_ID) %>%
  mutate(Match_flag = ifelse(MatchFactor < 700,
                             "Low score (<700)", "")) %>%
  ungroup()

```

**NOTE.** This step was used to quality control, the result of this was verify by manual identification.

Deleiting hits with low match.

```{r }

mn_clean <- mn_flag %>%
  group_by(Feature_ID) %>%
  group_modify(~{
    if (all(.x$Match_flag == "Low score (<700)")) {
      .x %>%
        slice(1) %>%
        mutate(across(-c(Segment, Alignment_ID, NIST_Code, RT, Exp_RI,
                         Match_flag), ~ NA))
    } else {
      .x %>%
        filter(Match_flag != "Low score (<700)")
    }
  }) %>%
  ungroup()
# Orderening by the Feature_ID
mn_clean <- mn_clean %>%
  mutate(Feature_ID = as.numeric(Feature_ID)) %>%  # Change to numeric
  arrange(Feature_ID)
# Deleiting ionization tag in in-house library "[M]+.  70eV"
mn_clean <- mn_clean %>%
  mutate(
    Name = str_replace(Name, "\\s*\\[M\\]\\+\\.\\s*70\\s*eV", ""))

```

Now, we focus on feature with match with difference between experimental RI and literature RI between 0 and 10. In the case that the feature meet this condition, only match with high match score will be reported. If you have a in house library made using autentic reference standard you can prioritize this library in order to report level 1 identification level.

Now we focus in level 3 identification level. Here we separate in two filter, one features between 10 and 100 of RI desviation, and feature that have more of 100 RI desviation or its do not have RI desviation. In the fist case, we prioritize the low RI desviation, and in the second case we prioritaze the high match factor.

```{r }

mn_il <- mn_clean %>%
  group_by(Feature_ID) %>%
  mutate(
    # Clasificación por rangos
    in_0_10   = !is.na(RI_Error) & RI_Error <= 10,
    in_10_100 = !is.na(RI_Error) & RI_Error > 10 & RI_Error <= 100,
    has_0_10   = any(in_0_10),
    has_10_100 = any(in_10_100),
    is_ikiam = Library == "ikiam_npl_gcms_lib") %>%
  # Asignar prioridad por fila
  mutate(priority = case_when(
      has_0_10   & in_0_10   ~ 1,
      !has_0_10 & has_10_100 & in_10_100 ~ 2,
      !has_0_10 & !has_10_100 ~ 3,
      TRUE ~ NA_real_)) %>%
  # Mantener solo filas relevantes por feature
  filter(!is.na(priority)) %>%
  # Orden según prioridad
  arrange(priority,
    # PRIORIDAD 1 → Match alto primero o in-house library si aplica
    desc(ifelse(priority == 1, is_ikiam, FALSE)),
    desc(ifelse(priority == 1, MatchFactor, NA)),
    ifelse(priority == 1, RI_Error, NA),
    # PRIORIDAD 2 → RI bajo primero
    ifelse(priority == 2, RI_Error, NA),
    desc(ifelse(priority == 2, MatchFactor, NA)),
    # PRIORIDAD 3 → Match alto
    desc(ifelse(priority == 3, MatchFactor, NA))) %>%
  slice(1) %>%
  ungroup() %>%
  select(-in_0_10, -in_10_100, -has_0_10, -has_10_100, -priority, -is_ikiam)
# Orderening by the Feature_ID
mn_il <- mn_il %>%
  mutate(Feature_ID = as.numeric(Feature_ID)) %>%  # Change to numeric
  arrange(Feature_ID)

```

En some case same name of metabolite has been asigned for multiple feature. In the case with same metabolite name was assigned for multiple feature, the best match factor was priorized for features between 0 to 10 RI error (the in hose library was priorized a them the filter was apply). For features with same metabolite name and RI error between 10 and 100, the metabolie name was assigned for feature with the low RI error, in the case with the same RI error the best match factor was priorized. For features with the same metabolite name and RI error > 100 or that have NA, the metabolite name was assigned to feature with major match factor. Feature that not meet the condition descrive above was tag as "Repeated (RI error <10)", "Repeated (RI error 10-100)", and "Repeated (RI error >100/NA)", accordingly.

```{r }

# Separar identificados
mn_ids <- mn_il %>%
  filter(!is.na(Name))
# Separar no identificados
mn_noids <- mn_il %>%
  filter(is.na(Name))
# Aplicar lógica solo a identificados
mn_rep <- mn_ids %>%   # Tabla con 1 hit por feature
  group_by(Name) %>%
  mutate(in_0_10   = !is.na(RI_Error) & RI_Error <= 10,
         in_10_100 = !is.na(RI_Error) & RI_Error > 10 & RI_Error <= 100,
         has_0_10   = any(in_0_10),
         has_10_100 = any(in_10_100),
         is_ikiam = Library == "ikiam_npl_gcms_lib") %>%
  # Definir grupo activo
  mutate(active_group = case_when(
    has_0_10 ~ "0_10",
    !has_0_10 & has_10_100 ~ "10_100",
    TRUE ~ "gt100")) %>%
  # Definir quién participa en el ranking
  mutate(candidate = case_when(active_group == "0_10"   ~ in_0_10,
                               active_group == "10_100" ~ in_10_100,
                               TRUE ~ TRUE)) %>%
  # Orden SOLO para candidatos
  arrange(Name,
    desc(candidate),  # candidatos primero
    # PRIORIDAD 1
    desc(ifelse(candidate & active_group == "0_10", is_ikiam, FALSE)),
    desc(ifelse(candidate & active_group == "0_10", MatchFactor, NA)),
    ifelse(candidate & active_group == "0_10", RI_Error, Inf),
    # PRIORIDAD 2
    ifelse(candidate & active_group == "10_100", RI_Error, Inf),
    desc(ifelse(candidate & active_group == "10_100", MatchFactor, NA)),
    # PRIORIDAD 3
    desc(ifelse(candidate & active_group == "gt100", MatchFactor, NA))) %>%
  group_by(Name) %>%
  mutate(rank_name = ifelse(candidate, row_number(), NA_integer_)) %>%
  ungroup() %>%
  # Etiquetado
  mutate(Match_flag = case_when(
    rank_name == 1 ~ Match_flag,
    active_group == "0_10" ~ "Repeated (RI error <10)",
    active_group == "10_100" ~ "Repeated (RI error 10-100)",
    TRUE ~ "Repeated (RI error >100/NA)")) %>%
  select(-in_0_10, -in_10_100, -has_0_10, -has_10_100,
         -is_ikiam, -active_group, -candidate, -rank_name)
# Volver a unir todo
mn_rep <- bind_rows(mn_rep, mn_noids)
# Orderening by the Feature_ID
mn_rep <- mn_rep %>%
  mutate(Feature_ID = as.numeric(Feature_ID)) %>%  # Change to numeric
  arrange(Feature_ID)

```

Adding a identification level, metabolites identified with in house library (match factor > 700) and RI error < 10, will be tag as level 1 identification confidence level, metabolites identified with NIST libraries (match factor > 700) and RI error < 10, wil be tag as level 2 identification confidence level. Metabolites identified wit NIST libraries (match factor > 700) and RI error > 10 was assigned as level 3 identification confidense level.

```{r }

mn_il123 <- mn_rep %>%
  mutate(IL = case_when(
      # No etiquetar
      is.na(Name) ~ NA_character_,
      Match_flag == "Low score (<700)" ~ NA_character_,
      # Level 1
      !is.na(RI_Error) & RI_Error <= 10 & 
        Library == "ikiam_npl_gcms_lib" ~ "Level 1",
      # Level 2
      !is.na(RI_Error) & RI_Error <= 10 ~ "Level 2",
      # Level 3
      TRUE ~ "Level 3"))

```

Extraction NIST 20 library mass spectrum.

```{r }

# Load the NIST database (spectrum information of library)
nist_mz <- readRDS("F:/NIST_20_Library/Result/nist_mz.rds")

```

Adding espectrum data from the NIST libry to the identified metabolites.

```{r }

mn_libmz <- mn_il123 %>%
  mutate(DB_ID = as.character(DB_ID)) %>%
  left_join(nist_mz %>%
              mutate(DB_ID = as.character(DB_ID)) %>%
              select(DB_ID, Library, nist_mz),
            by = c("DB_ID", "Library"))

```

Now we going to add the experimental EI spectrum to the identified feature, in the following steps we could compare the spectrum between NIST spectrum and experimental spectrum.

```{r }

# Funtion to extract experimental EI spectrum from *.mgf file exported from MZmine
parse_mzmine_mgf <- function(file_path) {
  lines <- readLines(file_path)
  data_list <- list()
  current_id <- NA
  current_rt <- NA
  mz_vec <- c()
  int_vec <- c()
  inside_block <- FALSE
  for (line in lines) {
    # Begining of block
    if (str_detect(line, "^BEGIN IONS")) {
      inside_block <- TRUE
      current_id <- NA
      current_rt <- NA
      mz_vec <- c()
      int_vec <- c()
    }
    if (inside_block) {
      # Feature ID
      if (str_detect(line, "^FEATURE_ID=")) {
        current_id <- as.integer(str_extract(line, "\\d+"))
      }
      # RT
      if (str_detect(line, "^RTINSECONDS=")) {
        current_rt <- as.numeric(str_extract(line, "\\d+\\.?\\d*"))
      }
      # Row of spectrum (m/z intensidad)
      if (str_detect(line, "^[0-9]")) {
        parts <- str_split(line, "\\s+")[[1]]
        mz_vec <- c(mz_vec, as.numeric(parts[1]))
        int_vec <- c(int_vec, as.numeric(parts[2]))
      }
    }
    # End of block
    if (str_detect(line, "^END IONS")) {
      row <- tibble::tibble(
        Alignment_ID = current_id,
        RT_sec = current_rt,
        Exp_mz = list(mz_vec),
        Exp_int = list(int_vec)
      )
      data_list[[length(data_list) + 1]] <- row
      inside_block <- FALSE
    }
  }
  dplyr::bind_rows(data_list)
}
# Reading *.mgf files (MZmine EI dec results)
mgf_file_mn3t49m <- "../X_hylaeae_metabolomics/Data/MZmine_GCMS_apolar_3-49min_to_NIST.mgf"
mgf_file_mn49t54m <- "../X_hylaeae_metabolomics/Data/MZmine_GCMS_apolar_49-54min_to_NIST.mgf"
mgf_file_mn54t56m <- "../X_hylaeae_metabolomics/Data/MZmine_GCMS_apolar_54-56min_to_NIST.mgf"
# Apply for each segment
mgf_mn3t49m_df  <- parse_mzmine_mgf(mgf_file_mn3t49m)
mgf_mn49t54m_df <- parse_mzmine_mgf(mgf_file_mn49t54m)
mgf_mn54t56m_df <- parse_mzmine_mgf(mgf_file_mn54t56m)
# Merge all
mn_ei <- bind_rows(
  mgf_mn3t49m_df %>% mutate(Segment = "3-49min"),
  mgf_mn49t54m_df %>% mutate(Segment = "49-54min"),
  mgf_mn54t56m_df %>% mutate(Segment = "54-56min"))
# Addin experimental spectum to identified metabolites
mn_expmz <- mn_libmz %>%
  left_join(mn_ei %>%
              select(Alignment_ID, Segment, Exp_mz, Exp_int),
            by = c("Alignment_ID", "Segment"))
# Adding intesity filter (nois)
mn_expmz_ns <- mn_expmz %>%
  rowwise() %>%
  mutate(int_exp_norm = list(Exp_int / max(Exp_int, na.rm = TRUE) * 100),
         keep_idx = list(which(int_exp_norm >= 0)), # IMPORTANT use the same noise used in NIST MS Search
         exp_mz_ns = list(round(Exp_mz[keep_idx])),
         exp_int_ns = list(int_exp_norm[keep_idx])) %>%
  ungroup()

```

The folowing code count the number of ions are equal between library and experiment spectrum.

```{r }

# Compare to only identified features
mn_expmz_id <- mn_expmz_ns %>%
  filter(!is.na(Name)) %>%
  rowwise() %>%
  mutate(matched_ions = length(intersect(exp_mz_ns, nist_mz))) %>%
  ungroup()
# Extracting no identified features
mn_expmz_noid <- mn_expmz_ns %>%
  filter(is.na(Name)) %>%
  mutate(matched_ions = NA_integer_)
# Join identified and no identified features with number of ion match
mn_ions <- bind_rows(mn_expmz_id, mn_expmz_noid)
# Orderening by the Feature_ID
mn_ions <- mn_ions %>%
  mutate(Feature_ID = as.numeric(Feature_ID)) %>%  # Change to numeric
  arrange(Feature_ID)

```

Adding a flag if the number of ions match is below of 5.

```{r }

mn_ions_fg <- mn_ions %>%
  mutate(Match_flag = case_when(
    # No identificado (no editar)
    is.na(Name) ~ Match_flag,
    # Match de iones bajo
    !is.na(matched_ions) & matched_ions < 5 ~ 
      ifelse(is.na(Match_flag), "Low ion match (<5)",
             paste(Match_flag, "Low ion match (<5)", sep = "; ")),
    # Mantener lo demás
    TRUE ~ Match_flag))

```

Flag features identified as siloxane compounds by compund formula.

```{r }

mn_ions_si <- mn_ions_fg %>%
  mutate(Match_flag = case_when(
    !is.na(Formula) & str_detect(Formula, "Si\\d*") ~
      ifelse(is.na(Match_flag),
             "Siloxane",
             paste(Match_flag, "Siloxane", sep = "; ")),
    TRUE ~ Match_flag))

```

Extracting InChIKey for classification of type of molecule using a Classyfire tool (https://github.com/aberHRML/classyfireR).

```{r }

# Extractiong the InChIKey from the feature file
mn_ions_ik <- mn_ions_si %>%
  group_by(InChIKey) %>%
  filter(any(is.na(Match_flag) | Match_flag == "")) %>%
  ungroup() %>%
  filter(!is.na(InChIKey)) %>%
  distinct(InChIKey)
# Metabolite classification
mn_class <- purrr::map_dfr(mn_ions_ik$InChIKey, function(ik) {
  res <- tryCatch(get_classification(ik), error = function(e) NULL)
  if (is.null(res) || is.null(res@classification)) {
    return(tibble(
      InChIKey = ik,
      Superclass = NA,
      Class = NA,
      Subclass = NA
    ))
  }
  class_vec <- res@classification$Classification
  # Validar vector
  if (!is.character(class_vec) || length(class_vec) < 2) {
    return(tibble(
      InChIKey = ik,
      Superclass = NA,
      Class = NA,
      Subclass = NA
    ))
  }
  tibble(
    InChIKey = ik,
    Superclass = ifelse(length(class_vec) >= 2, class_vec[2], NA),
    Class      = ifelse(length(class_vec) >= 3, class_vec[3], NA),
    Subclass   = ifelse(length(class_vec) >= 4, class_vec[4], NA)
  )
})

```

Adding a classification of moleculecules to the feature table of identified features.

```{r }

# Unir con tabla original the molecules classification
mn_ions_cf <- mn_ions_si %>%
  left_join(mn_class, by = "InChIKey")
# Exporting the feature list with metabolites identified al IL 1, 2, and 3
mn_ions_cf %>%
  select(Feature_ID, Segment, Alignment_ID, IL, RT, Name, Synon1, Synon2, Synon3,
         Formula, MatchFactor, matched_ions, Lit_RI, Exp_RI, RI_Error, Library,
         DB_ID, InChIKey, Superclass, Class, Subclass, Match_flag, RI_tag,
         nist_mz, exp_mz_ns, exp_int_ns) %>%
  mutate(nist_mz = sapply(nist_mz, paste, collapse = ", "),
         exp_mz_ns = sapply(exp_mz_ns, paste, collapse = ", "),
         exp_int_ns = sapply(exp_int_ns, paste, collapse = ", ")) %>%
  write_xlsx("../X_hylaeae_metabolomics/Result/MZmine_GCMS_apolar_IDs.xlsx")

```

